---
title: "DreamAI imputation validation and DEqMS analysis"
author: "Isabella Pauser"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    self_contained: true
    toc: true
    toc_float: true
params:
  validation_runs: 10
---

# Analysis overview

This report performs the DreamAI/DEqMS proteomics workflow and evaluates the
imputation step by repeatedly masking observed values. Because the true values
at naturally missing positions are unknown, validation is performed on complete
proteins. Two masking schemes are considered: uniformly random missingness
(MCAR) and low-abundance-weighted missingness (MNAR-style).

Imputation performance is assessed using the correlation between observed and
imputed intensities (ACC_OI), normalized root-mean-square error (NRMSE), and the
Procrustes discrepancy between PCA score spaces (PSS). Higher ACC_OI and lower
NRMSE and PSS values indicate better agreement. SOR summarizes the ranks of the
sample-specific NRMSE values and is interpreted only relative to the methods
included in this analysis.

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}

project_dir <- "~/Proteomics_Melanoma_Pauser"
setwd(project_dir)

# -----------------------------
# Load packages
# -----------------------------

library(DEqMS)
library(limma)
library(matrixStats)
library(ggplot2)
library(ggrepel)
library(DreamAI)
library(naniar)
library(dplyr)
library(tibble)
library(pheatmap)
library(gridExtra)
library(grid)

# -----------------------------
# Paths
# -----------------------------

input_file <- "proteinGroups.txt"

output_dir <- "Proteomics_DreamAI_results"

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

```

## 2. Import the MaxQuant data

The MaxQuant `proteinGroups.txt` table is imported. Reverse hits and potential
contaminants are removed.

```{r import-maxquant}

df.prot <- read.table(
  input_file,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  comment.char = "",
  quote = "",
  check.names = TRUE
)

# Inspect contaminant/reverse columns
grep("Reverse|contaminant|Contaminant", names(df.prot), value = TRUE)

# Remove reverse hits and contaminants
df.prot <- df.prot[
  (is.na(df.prot$Reverse) | df.prot$Reverse != "+") &
    (is.na(df.prot$Potential.contaminant) | df.prot$Potential.contaminant != "+"),
]

rownames(df.prot) <- df.prot$Majority.protein.IDs

dim(df.prot)

```

## 3. Select and filter LFQ measurements

LFQ columns for MCM-1G and MCM-DLN are selected, zeros are converted to NA,
and proteins must contain at least two quantified values in each condition.

```{r filter-lfq}

lfq_cols <- grep(
  "^LFQ\\.intensity\\.(1G_A|1G_B|1G_C|DLN_A|DLN_B|DLN_C)_",
  names(df.prot),
  value = TRUE
)

length(lfq_cols)
lfq_cols

df.LFQ <- df.prot[, lfq_cols]

# MaxQuant LFQ values of 0 are treated as missing values
df.LFQ[df.LFQ == 0] <- NA
rownames(df.LFQ) <- df.prot$Majority.protein.IDs

# -----------------------------
# Filter proteins
# Require at least 2 valid LFQ values per condition
# -----------------------------

cols_1G <- grep("^LFQ\\.intensity\\.1G_", names(df.LFQ))
cols_DLN <- grep("^LFQ\\.intensity\\.DLN_", names(df.LFQ))

df.LFQ$na_count_1G <- apply(df.LFQ[, cols_1G], 1, function(x) sum(is.na(x)))
df.LFQ$na_count_DLN <- apply(df.LFQ[, cols_DLN], 1, function(x) sum(is.na(x)))

df.LFQ.filter <- df.LFQ[
  df.LFQ$na_count_1G <= length(cols_1G) - 2 &
    df.LFQ$na_count_DLN <= length(cols_DLN) - 2,
  c(cols_1G, cols_DLN)
]

dim(df.LFQ.filter)

```

## 4. Prepare peptide counts and the log2 LFQ matrix

Peptide counts are retained for DEqMS variance modelling. The filtered LFQ
intensities are converted to a protein-by-sample log2 matrix.

```{r prepare-analysis-matrices}

pep_cols <- grep(
  "^Peptides\\.(1G_A|1G_B|1G_C|DLN_A|DLN_B|DLN_C)_",
  names(df.prot),
  value = TRUE
)

length(pep_cols)
pep_cols

pep.count.table <- data.frame(
  count = rowMins(as.matrix(df.prot[, pep_cols]), na.rm = TRUE),
  row.names = df.prot$Majority.protein.IDs
)

# DEqMS requires positive peptide counts
# DEqMS requires a positive count; preserve valid counts and repair only zeros.
pep.count.table$count[pep.count.table$count < 1] <- 1

summary(pep.count.table$count)

# -----------------------------
# Prepare log2 LFQ matrix
# -----------------------------

lfq.matrix <- as.matrix(df.LFQ.filter)
log2.lfq <- log2(lfq.matrix)

dim(log2.lfq)

```

## 5. Assess missing values before imputation

```{r missing-value-assessment, fig.width=11, fig.height=7}

missing.data <- as.data.frame(log2.lfq)

colnames(missing.data) <- gsub(
  "LFQ\\.intensity\\.",
  "",
  colnames(missing.data)
)

missing.data <- missing.data[
  c(
    grep("^1G_", colnames(missing.data), value = TRUE),
    grep("^DLN_", colnames(missing.data), value = TRUE)
  )
]

missing.plot <- vis_miss(missing.data) +
  ggtitle("Missing values before DreamAI: MCM-1G vs MCM-DLN") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )

missing.plot

ggsave(
  filename = file.path(output_dir, "Missing_Data_Pattern_Before_DreamAI_Grouped.jpg"),
  plot = missing.plot,
  width = 12,
  height = 7,
  dpi = 600
)

# -----------------------------
# Missing value summary
# -----------------------------

missing_per_sample <- colSums(is.na(log2.lfq))
missing_percent_sample <- colMeans(is.na(log2.lfq)) * 100
total_missing <- sum(is.na(log2.lfq))
overall_missing_percent <- mean(is.na(log2.lfq)) * 100

missing.summary <- data.frame(
  Sample = colnames(log2.lfq),
  Missing_values = missing_per_sample,
  Missing_percent = round(missing_percent_sample, 2)
)

overall.missing.summary <- data.frame(
  Total_missing_values = total_missing,
  Overall_missing_percent = round(overall_missing_percent, 2)
)

missing.summary
overall.missing.summary

write.table(
  missing.summary,
  file.path(output_dir, "Missing_values_per_sample_before_DreamAI.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

write.table(
  overall.missing.summary,
  file.path(output_dir, "Overall_missing_values_before_DreamAI.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```

## 6. Impute missing values with DreamAI

DreamAI combines KNN, MissForest, ADMIN, Birnn, SpectroFM and RegImpute in an
ensemble. The resulting matrix is checked for its dimensions and remaining
missing values.

```{r dreamai-imputation}

dreamAI.out <- DreamAI(
  data = log2.lfq,
  k = 10,
  maxiter_MF = 10,
  ntree = 100,
  maxnodes = NULL,
  maxiter_ADMIN = 30,
  tol = 10^(-2),
  gamma_ADMIN = NA,
  gamma = 50,
  CV = FALSE,
  fillmethod = "row_mean",
  maxiter_RegImpute = 10,
  conv_nrmse = 1e-06,
  iter_SpectroFM = 40,
  method = c(
    "KNN",
    "MissForest",
    "ADMIN",
    "Birnn",
    "SpectroFM",
    "RegImpute"
  ),
  out = c("Ensemble")
)

str(dreamAI.out)
names(dreamAI.out)

protein.matrix <- dreamAI.out$Ensemble

```

## 7. Validate DreamAI by artificial masking

### 7.1 Validation settings and helper functions

```{r validation-functions}

validation_runs <- as.integer(params$validation_runs)
validation_scenarios <- c("MAR", "MNAR-style")
validation_mask_fraction <- mean(is.na(log2.lfq))
validation_reference <- log2.lfq[complete.cases(log2.lfq), , drop = FALSE]

if (nrow(validation_reference) < 10L) {
  stop("DreamAI validation requires at least 10 proteins with complete observations.")
}
if (!is.finite(validation_mask_fraction) || validation_mask_fraction <= 0) {
  stop("The original matrix has no missing values; artificial masking fraction is undefined.")
}

run_dreamai_ensemble <- function(x) {
  fit <- DreamAI(
    data = x, k = 10, maxiter_MF = 10, ntree = 100, maxnodes = NULL,
    maxiter_ADMIN = 30, tol = 10^(-2), gamma_ADMIN = NA, gamma = 50,
    CV = FALSE, fillmethod = "row_mean", maxiter_RegImpute = 10,
    conv_nrmse = 1e-06, iter_SpectroFM = 40,
    method = c("KNN", "MissForest", "ADMIN", "Birnn", "SpectroFM", "RegImpute"),
    out = "Ensemble"
  )
  as.matrix(fit$Ensemble)
}

make_validation_mask <- function(reference, fraction, scenario, max_attempts = 100L) {
  n_mask <- max(2L, min(nrow(reference) * ncol(reference) - 1L,
                        floor(fraction * length(reference))))
  values <- as.numeric(reference)
  weights <- if (scenario == "MAR") {
    rep(1, length(values))
  } else {
    z <- as.numeric(scale(values))
    plogis(-z) + 1e-06
  }

  for (attempt in seq_len(max_attempts)) {
    selected <- sample.int(length(values), n_mask, replace = FALSE, prob = weights)
    mask <- matrix(FALSE, nrow(reference), ncol(reference),
                   dimnames = dimnames(reference))
    mask[selected] <- TRUE
    retained_by_row <- rowSums(!mask)
    retained_by_col <- colSums(!mask)
    if (all(retained_by_row >= 2L) && all(retained_by_col >= 2L)) return(mask)
  }
  stop("Could not construct a valid artificial mask after ", max_attempts, " attempts.")
}

fill_reference_methods <- function(masked) {
  row_fill <- masked
  row_means <- rowMeans(masked, na.rm = TRUE)
  row_fill[is.na(row_fill)] <- row_means[row(row_fill)[is.na(row_fill)]]

  column_fill <- masked
  column_means <- colMeans(masked, na.rm = TRUE)
  column_fill[is.na(column_fill)] <- column_means[col(column_fill)[is.na(column_fill)]]

  minimum_fill <- masked
  minimum_fill[is.na(minimum_fill)] <- min(masked, na.rm = TRUE)

  list(`Row mean` = row_fill, `Column mean` = column_fill,
       `Global minimum` = minimum_fill)
}

calculate_pss <- function(reference, reconstructed, n_components = 5L) {
  x <- prcomp(t(reference), center = TRUE, scale. = FALSE)$x
  y <- prcomp(t(reconstructed), center = TRUE, scale. = FALSE)$x
  k <- min(n_components, ncol(x), ncol(y), nrow(x) - 1L)
  x <- scale(x[, seq_len(k), drop = FALSE], center = TRUE, scale = FALSE)
  y <- scale(y[, seq_len(k), drop = FALSE], center = TRUE, scale = FALSE)
  x <- x / sqrt(sum(x^2))
  y <- y / sqrt(sum(y^2))
  rotation_svd <- svd(t(y) %*% x)
  rotation <- rotation_svd$u %*% t(rotation_svd$v)
  sum((x - y %*% rotation)^2)
}

calculate_metrics <- function(truth, prediction) {
  data.frame(
    ACC_OI = cor(truth, prediction, use = "complete.obs"),
    NRMSE = sqrt(mean((truth - prediction)^2) / var(truth))
  )
}

```

### 7.2 Run repeated validation

```{r run-validation}
metric_results <- list()
prediction_results <- list()
sample_nrmse_results <- list()
validation_status <- list()
result_index <- 0L

set.seed(20260801)
for (scenario in validation_scenarios) {
  for (run_id in seq_len(validation_runs)) {
    message("DreamAI validation: ", scenario, ", repetition ", run_id,
            " of ", validation_runs)
    result_index <- result_index + 1L

    run_result <- tryCatch({
      artificial_mask <- make_validation_mask(
        validation_reference, validation_mask_fraction, scenario
      )
      masked_matrix <- validation_reference
      masked_matrix[artificial_mask] <- NA_real_

      method_matrices <- c(
        list(DreamAI = run_dreamai_ensemble(masked_matrix)),
        fill_reference_methods(masked_matrix)
      )

      run_metrics <- list()
      run_predictions <- list()
      run_sample_nrmse <- list()

      for (method_name in names(method_matrices)) {
        completed <- method_matrices[[method_name]]
        reconstructed <- validation_reference
        reconstructed[artificial_mask] <- completed[artificial_mask]
        truth <- validation_reference[artificial_mask]
        prediction <- completed[artificial_mask]

        global_metrics <- calculate_metrics(truth, prediction)
        global_metrics$PSS <- calculate_pss(validation_reference, reconstructed)
        global_metrics$scenario <- scenario
        global_metrics$run <- run_id
        global_metrics$method <- method_name
        run_metrics[[method_name]] <- global_metrics

        mask_locations <- which(artificial_mask, arr.ind = TRUE)
        run_predictions[[method_name]] <- data.frame(
          scenario = scenario, run = run_id, method = method_name,
          protein = rownames(validation_reference)[mask_locations[, 1]],
          sample = colnames(validation_reference)[mask_locations[, 2]],
          truth = truth, prediction = prediction,
          error = prediction - truth
        )

        per_sample <- lapply(seq_len(ncol(validation_reference)), function(j) {
          selected <- artificial_mask[, j]
          sample_truth <- validation_reference[selected, j]
          sample_prediction <- completed[selected, j]
          value <- if (length(sample_truth) >= 2L && var(sample_truth) > 0) {
            sqrt(mean((sample_truth - sample_prediction)^2) / var(sample_truth))
          } else NA_real_
          data.frame(scenario = scenario, run = run_id, method = method_name,
                     sample = colnames(validation_reference)[j], NRMSE = value)
        })
        run_sample_nrmse[[method_name]] <- do.call(rbind, per_sample)
      }

      list(
        metrics = do.call(rbind, run_metrics),
        predictions = do.call(rbind, run_predictions),
        sample_nrmse = do.call(rbind, run_sample_nrmse)
      )
    }, error = function(e) e)

    if (inherits(run_result, "error")) {
      validation_status[[result_index]] <- data.frame(
        scenario = scenario, run = run_id, status = "failed",
        detail = conditionMessage(run_result)
      )
    } else {
      metric_results[[result_index]] <- run_result$metrics
      prediction_results[[result_index]] <- run_result$predictions
      sample_nrmse_results[[result_index]] <- run_result$sample_nrmse
      validation_status[[result_index]] <- data.frame(
        scenario = scenario, run = run_id, status = "completed", detail = NA_character_
      )
    }
  }
}

```

### 7.3 Combine validation results and calculate SOR

```{r validation-results}
validation_status_table <- do.call(rbind, validation_status)
if (length(metric_results) == 0L) {
  stop("All validation repetitions failed. Inspect validation_status_table.")
}

validation_metrics <- do.call(rbind, metric_results)
validation_predictions <- do.call(rbind, prediction_results)
validation_sample_nrmse <- do.call(rbind, sample_nrmse_results)

validation_sor <- validation_sample_nrmse %>%
  group_by(scenario, run, sample) %>%
  mutate(rank = rank(NRMSE, ties.method = "average", na.last = "keep")) %>%
  ungroup() %>%
  group_by(scenario, run, method) %>%
  summarise(SOR = sum(rank, na.rm = TRUE), .groups = "drop")

validation_metric_long <- validation_metrics %>%
  dplyr::select(
    scenario,
    run,
    method,
    ACC_OI,
    NRMSE,
    PSS
  ) %>%
  tidyr::pivot_longer(
    cols = c(ACC_OI, NRMSE, PSS),
    names_to = "metric",
    values_to = "value"
  )
```

### 7.4 Create validation diagnostic plots

```{r validation-plots, fig.width=12, fig.height=8}

metric_plot <- ggplot(validation_metric_long,
                      aes(x = method, y = value, fill = method)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, alpha = 0.6, size = 1.2) +
  facet_grid(metric ~ scenario, scales = "free_y") +
  labs(title = "Artificial-masking imputation validation",
       subtitle = "ACC_OI: higher is better; NRMSE and PSS: lower are better",
       x = NULL, y = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "none")

sor_plot <- ggplot(validation_sor, aes(x = method, y = SOR, fill = method)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, alpha = 0.6, size = 1.2) +
  facet_wrap(~scenario, scales = "free_y") +
  labs(title = "Sum of per-sample NRMSE ranks (SOR)",
       subtitle = "Lower is better within this reference-method set",
       x = NULL, y = "SOR") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "none")

dreamai_predictions <- validation_predictions %>% filter(method == "DreamAI")
scatter_plot <- ggplot(dreamai_predictions,
                       aes(x = truth, y = prediction, colour = scenario)) +
  geom_point(alpha = 0.18, size = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~scenario) +
  coord_equal() +
  labs(title = "DreamAI: held-out versus imputed intensities",
       x = "Observed log2 LFQ intensity", y = "Imputed log2 LFQ intensity") +
  theme_bw() + theme(legend.position = "none")

error_plot <- ggplot(dreamai_predictions,
                     aes(x = truth, y = abs(error), colour = scenario)) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~scenario) +
  labs(title = "DreamAI absolute error across abundance",
       x = "Observed log2 LFQ intensity", y = "Absolute imputation error") +
  theme_bw() + theme(legend.position = "none")

print(metric_plot)
print(sor_plot)
print(scatter_plot)
print(error_plot)
```

### 7.5 Save validation results

```{r save-validation-results}

write.table(validation_metrics,
            file.path(output_dir, "DreamAI_validation_metrics.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(validation_sor,
            file.path(output_dir, "DreamAI_validation_SOR.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(validation_predictions,
            file.path(output_dir, "DreamAI_validation_predictions.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(validation_status_table,
            file.path(output_dir, "DreamAI_validation_status.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
saveRDS(list(metrics = validation_metrics, SOR = validation_sor,
             predictions = validation_predictions,
             status = validation_status_table),
        file.path(output_dir, "DreamAI_validation_results.rds"))

ggsave(file.path(output_dir, "DreamAI_validation_metrics.jpg"), metric_plot,
       width = 12, height = 9, units = "in", dpi = 600)
ggsave(file.path(output_dir, "DreamAI_validation_SOR.jpg"), sor_plot,
       width = 10, height = 6, units = "in", dpi = 600)
ggsave(file.path(output_dir, "DreamAI_validation_scatter.jpg"), scatter_plot,
       width = 10, height = 5, units = "in", dpi = 600)
ggsave(file.path(output_dir, "DreamAI_validation_error_by_abundance.jpg"), error_plot,
       width = 10, height = 5, units = "in", dpi = 600)

dim(protein.matrix)
sum(is.na(protein.matrix))

```

## 8. Save the DreamAI matrix and inspect distributions

```{r dreamai-distribution-qc, fig.width=11, fig.height=7}
# Save DreamAI matrix
write.table(
  protein.matrix,
  file.path(output_dir, "DreamAI_imputed_Melanoma_Ensemble.tsv"),
  sep = "\t",
  quote = FALSE
)

# -----------------------------
# Density plot before and after DreamAI
# -----------------------------

density.df <- rbind(
  data.frame(
    Intensity = as.numeric(log2.lfq),
    Status = "Original observed LFQ"
  ),
  data.frame(
    Intensity = as.numeric(protein.matrix),
    Status = "DreamAI complete LFQ"
  )
)

density.df <- density.df[is.finite(density.df$Intensity), ]

density.plot <- ggplot(
  density.df,
  aes(x = Intensity, fill = Status)
) +
  geom_density(alpha = 0.45) +
  theme_bw(base_size = 14) +
  labs(
    title = "Density plot of LFQ intensities: original vs DreamAI",
    x = "log2 LFQ intensity",
    y = "Density"
  )

density.plot

ggsave(
  filename = file.path(output_dir, "DreamAI_DensityPlot_Original_vs_Imputed.jpg"),
  plot = density.plot,
  width = 7,
  height = 5,
  dpi = 600
)

# -----------------------------
# Boxplot before and after DreamAI
# -----------------------------

jpeg(
  file.path(output_dir, "DreamAI_Boxplot_Before_After.jpg"),
  width = 8000,
  height = 4000,
  res = 600,
  quality = 100
)

par(mfrow = c(1, 2), mar = c(10, 4, 4, 2))

boxplot(
  log2.lfq,
  las = 2,
  main = "Before DreamAI",
  ylab = "log2 LFQ intensity"
)

boxplot(
  protein.matrix,
  las = 2,
  main = "After DreamAI",
  ylab = "log2 LFQ intensity"
)

par(mfrow = c(1, 1))
dev.off()

# -----------------------------
# Final sample distribution QC
# -----------------------------

jpeg(
  file.path(output_dir, "Sample_distribution_QC.jpg"),
  width = 5000,
  height = 4500,
  res = 600,
  quality = 100
)

boxplot(
  protein.matrix,
  las = 2,
  main = "Sample LFQ intensity distributions",
  ylab = "log2 LFQ intensity"
)

dev.off()

```

## 9. PCA before and after imputation

```{r pca-qc, fig.width=10, fig.height=7}

pca.before.matrix <- log2.lfq[complete.cases(log2.lfq), ]

pca.before <- prcomp(
  t(pca.before.matrix),
  scale = TRUE
)

pca.after <- prcomp(
  t(protein.matrix),
  scale = TRUE
)

print(pca.before)
summary(pca.before)

print(pca.after)
summary(pca.after)

# -----------------------------
# Sample metadata for PCA
# -----------------------------

make_sample_metadata <- function(sample.names) {
  
  sample.short <- gsub(
    "LFQ\\.intensity\\.",
    "",
    sample.names
  )
  
  condition <- ifelse(
    grepl("^1G_", sample.short),
    "MCM_1G",
    "MCM_DLN"
  )
  
  biological.replicate <- sub(
    ".*_([ABC])_.*",
    "\\1",
    sample.short
  )
  
  technical.replicate <- sub(
    ".*_([0-9]+)$",
    "\\1",
    sample.short
  )
  
  data.frame(
    Sample = sample.names,
    SampleShort = sample.short,
    Condition = factor(condition, levels = c("MCM_1G", "MCM_DLN")),
    BiologicalReplicate = factor(biological.replicate, levels = c("A", "B", "C")),
    TechnicalReplicate = technical.replicate,
    stringsAsFactors = FALSE
  )
}

# -----------------------------
# PCA variance plots
# -----------------------------

jpeg(
  file.path(output_dir, "PCA_variance_before_DreamAI.jpg"),
  width = 5000,
  height = 4000,
  res = 600,
  quality = 100
)

plot(
  pca.before,
  main = "PCA variance before DreamAI"
)

dev.off()

jpeg(
  file.path(output_dir, "PCA_variance_after_DreamAI.jpg"),
  width = 5000,
  height = 4000,
  res = 600,
  quality = 100
)

plot(
  pca.after,
  main = "PCA variance after DreamAI"
)

dev.off()

# -----------------------------
# PCA score plots
# -----------------------------

pca.before.meta <- make_sample_metadata(rownames(pca.before$x))
pca.after.meta <- make_sample_metadata(rownames(pca.after$x))

pca.before.var <- summary(pca.before)$importance[2, 1:2] * 100
pca.after.var <- summary(pca.after)$importance[2, 1:2] * 100

pca.before.df <- data.frame(
  PC1 = pca.before$x[, 1],
  PC2 = pca.before$x[, 2],
  pca.before.meta
)

pca.after.df <- data.frame(
  PC1 = pca.after$x[, 1],
  PC2 = pca.after$x[, 2],
  pca.after.meta
)

pca.before.plot <- ggplot(
  pca.before.df,
  aes(
    x = PC1,
    y = PC2,
    color = Condition,
    shape = BiologicalReplicate,
    label = SampleShort
  )
) +
  geom_point(size = 4) +
  geom_text_repel(size = 3, max.overlaps = Inf) +
  theme_bw(base_size = 14) +
  scale_shape_manual(
    values = c(
      "A" = 16,
      "B" = 17,
      "C" = 15
    )
  ) +
  ggtitle("PCA before DreamAI") +
  xlab(paste0("PC1 (", round(pca.before.var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca.before.var[2], 1), "%)")) +
  labs(
    color = "Condition",
    shape = "Biological replicate"
  )

ggsave(
  filename = file.path(output_dir, "PCA_before_DreamAI.jpg"),
  plot = pca.before.plot,
  width = 8,
  height = 6,
  dpi = 600
)

pca.after.plot <- ggplot(
  pca.after.df,
  aes(
    x = PC1,
    y = PC2,
    color = Condition,
    shape = BiologicalReplicate,
    label = SampleShort
  )
) +
  geom_point(size = 4) +
  geom_text_repel(size = 3, max.overlaps = Inf) +
  theme_bw(base_size = 14) +
  scale_shape_manual(
    values = c(
      "A" = 16,
      "B" = 17,
      "C" = 15
    )
  ) +
  ggtitle("PCA after DreamAI") +
  xlab(paste0("PC1 (", round(pca.after.var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca.after.var[2], 1), "%)")) +
  labs(
    color = "Condition",
    shape = "Biological replicate"
  )

ggsave(
  filename = file.path(output_dir, "PCA_after_DreamAI.jpg"),
  plot = pca.after.plot,
  width = 8,
  height = 6,
  dpi = 600
)

```

## 10. Differential protein abundance with DEqMS

The design contrasts MCM-1G against MCM-DLN. Limma fits the initial model and
DEqMS adjusts the empirical Bayes variance estimates using peptide counts.

```{r deqms-model}

class <- ifelse(
  grepl("^LFQ\\.intensity\\.1G_", colnames(protein.matrix)),
  "MCM_1G",
  "MCM_DLN"
)

class <- factor(class, levels = c("MCM_1G", "MCM_DLN"))

design <- model.matrix(~0 + class)
colnames(design) <- levels(class)

design

# -----------------------------
# Limma model
# -----------------------------

fit1 <- lmFit(protein.matrix, design = design)

cont <- makeContrasts(
  MCM_1G - MCM_DLN,
  levels = design
)

fit2 <- contrasts.fit(fit1, contrasts = cont)
fit3 <- eBayes(fit2)

# -----------------------------
# DEqMS model
# -----------------------------

fit3$count <- pep.count.table[
  rownames(fit3$coefficients),
  "count"
]

summary(fit3$count)
min(fit3$count, na.rm = TRUE)
sum(is.na(fit3$count))

stopifnot(!any(is.na(fit3$count)))
stopifnot(min(fit3$count, na.rm = TRUE) > 0)

fit4 <- spectraCounteBayes(
  fit3,
  fit.method = "loess",
  coef_col = 1
)

names(fit4)
head(fit4$sca.t)
head(fit4$sca.postvar)

# -----------------------------
# Extract DEqMS results
# -----------------------------

DEqMS.results <- outputResult(
  fit4,
  coef_col = 1
)

DEqMS.results$Gene.name <- df.prot[
  DEqMS.results$gene,
  "Gene.names"
]

# Final thesis cutoff
logFC_cutoff <- 1.0 # adjust to 0.58 for more sig. proteins
fdr_cutoff <- 0.05

DEqMS.results$significant <- with(
  DEqMS.results,
  sca.adj.pval <= fdr_cutoff & abs(logFC) > logFC_cutoff
)

head(DEqMS.results)
colnames(DEqMS.results)

write.table(
  DEqMS.results,
  file.path(output_dir, "DEqMS_results_MCM_1G_vs_MCM_DLN_DreamAI.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```

### 10.1 DEqMS diagnostic plots

```{r deqms-diagnostics, fig.width=9, fig.height=7}

jpeg(
  file.path(output_dir, "DEqMS_VarianceScatterplot.jpg"),
  width = 5000,
  height = 4500,
  res = 600,
  quality = 100
)

VarianceScatterplot(
  fit4,
  xlab = "log2(Peptide count + 1)",
  ylab = "log(Variance)",
  main = "MCM-1G vs MCM-DLN"
)

dev.off()

jpeg(
  file.path(output_dir, "DEqMS_Residualplot.jpg"),
  width = 5000,
  height = 4500,
  res = 600,
  quality = 100
)

Residualplot(
  fit4,
  xlab = "log2(Peptide count + 1)",
  ylab = "Variance(fitted - observed)",
  main = "MCM-1G vs MCM-DLN"
)

dev.off()

jpeg(
  file.path(output_dir, "DEqMS_VarianceBoxplot.jpg"),
  width = 5000,
  height = 4500,
  res = 600,
  quality = 100
)

VarianceBoxplot(
  fit4,
  n = max(fit4$count, na.rm = TRUE),
  xlab = "Peptide count + 1",
  ylab = "log(Variance)",
  main = "MCM-1G vs MCM-DLN"
)

dev.off()

```

### 10.2 Volcano and MA plots

Proteins are classified using an adjusted p-value of 0.05 and an absolute
log2-fold-change threshold of 1. Positive values indicate higher abundance in
MCM-1G; negative values indicate higher abundance in MCM-DLN.

```{r differential-abundance-plots, fig.width=10, fig.height=7}

DEqMS.results$log.sca.pval <- -log10(DEqMS.results$sca.adj.pval)

DEqMS.results$Regulation <- "Not significant"

DEqMS.results$Regulation[
  DEqMS.results$logFC > logFC_cutoff &
    DEqMS.results$sca.adj.pval <= fdr_cutoff
] <- "Up in MCM-1G"

DEqMS.results$Regulation[
  DEqMS.results$logFC < -logFC_cutoff &
    DEqMS.results$sca.adj.pval <= fdr_cutoff
] <- "Up in MCM-DLN"

DEqMS.results$Regulation <- factor(
  DEqMS.results$Regulation,
  levels = c("Up in MCM-DLN", "Up in MCM-1G", "Not significant")
)

top.labels <- DEqMS.results[
  DEqMS.results$Regulation != "Not significant" &
    !is.na(DEqMS.results$Gene.name) &
    DEqMS.results$Gene.name != "" &
    is.finite(DEqMS.results$log.sca.pval),
]

top.labels <- top.labels[order(top.labels$sca.adj.pval), ]
top.labels <- head(top.labels, 20)

volcano.plot <- ggplot(
  DEqMS.results,
  aes(x = logFC, y = log.sca.pval)
) +
  geom_point(aes(color = Regulation), size = 0.6) +
  theme_bw(base_size = 16) +
  xlab(expression("log2(MCM-1G / MCM-DLN)")) +
  ylab(expression("-log10(adjusted P-value)")) +
  geom_vline(xintercept = c(-logFC_cutoff, logFC_cutoff), colour = "red") +
  geom_hline(yintercept = -log10(fdr_cutoff), colour = "red") +
  geom_vline(xintercept = 0, colour = "black") +
  scale_colour_manual(
    values = c(
      "Up in MCM-DLN" = "blue",
      "Up in MCM-1G" = "orange",
      "Not significant" = "grey75"
    )
  ) +
  labs(
    colour = "Protein regulation",
    caption = paste0(
      "Cutoffs: |log2FC| > ",
      logFC_cutoff,
      " and DEqMS adjusted P-value <= ",
      fdr_cutoff
    )
  )

if (nrow(top.labels) > 0) {
  volcano.plot <- volcano.plot +
    geom_text_repel(
      data = top.labels,
      aes(label = Gene.name),
      max.overlaps = 20
    )
}

ggsave(
  filename = file.path(output_dir, "Volcano_plot_DreamAI_DEqMS.jpg"),
  plot = volcano.plot,
  width = 8,
  height = 7,
  dpi = 600
)

# -----------------------------
# MA plot
# -----------------------------

top.ma.labels <- DEqMS.results[
  abs(DEqMS.results$logFC) > logFC_cutoff &
    DEqMS.results$sca.adj.pval <= fdr_cutoff &
    !is.na(DEqMS.results$Gene.name) &
    DEqMS.results$Gene.name != "",
]

top.ma.labels <- top.ma.labels[
  order(top.ma.labels$sca.adj.pval),
]

top.ma.labels <- head(top.ma.labels, 20)

ma.plot <- ggplot(
  DEqMS.results,
  aes(
    x = AveExpr,
    y = logFC,
    color = Regulation
  )
) +
  geom_point(size = 0.6) +
  geom_hline(
    yintercept = c(-logFC_cutoff, 0, logFC_cutoff),
    color = c("red", "black", "red")
  ) +
  geom_text_repel(
    data = top.ma.labels,
    aes(label = Gene.name),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.3,
    point.padding = 0.2,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 14) +
  xlab("Average log2 LFQ intensity") +
  ylab(expression("log2(MCM-1G / MCM-DLN)")) +
  ggtitle("MA plot: MCM-1G vs MCM-DLN") +
  scale_colour_manual(
    values = c(
      "Up in MCM-DLN" = "blue",
      "Up in MCM-1G" = "orange",
      "Not significant" = "grey75"
    )
  )

ggsave(
  filename = file.path(output_dir, "MA_plot_DreamAI_DEqMS.jpg"),
  plot = ma.plot,
  width = 7,
  height = 6,
  dpi = 600
)


```

### 10.3 Heatmaps and individual protein plots

Proteins passing the predefined fold-change and adjusted-p-value thresholds are
displayed across samples. The matrix is standardized by protein before plotting.

```{r protein-heatmaps, fig.width=11, fig.height=8}

sig.proteins <- DEqMS.results[
  DEqMS.results$sca.adj.pval <= fdr_cutoff &
    abs(DEqMS.results$logFC) > logFC_cutoff,
]

nrow(sig.proteins)

sig.ids <- sig.proteins$gene
sig.ids <- sig.ids[sig.ids %in% rownames(protein.matrix)]

heatmap.matrix <- protein.matrix[sig.ids, , drop = FALSE]

dim(heatmap.matrix)

if (nrow(heatmap.matrix) < 2) {
  message("Only ", nrow(heatmap.matrix), 
          " significant protein found. A clustered heatmap is not meaningful.")
  
  write.table(
    sig.proteins,
    file.path(output_dir, "Significant_proteins_for_heatmap.txt"),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
} else {
  
  gene.labels <- DEqMS.results$Gene.name[
    match(rownames(heatmap.matrix), DEqMS.results$gene)
  ]
  
  gene.labels[is.na(gene.labels) | gene.labels == ""] <- rownames(heatmap.matrix)[
    is.na(gene.labels) | gene.labels == ""
  ]
  
  rownames(heatmap.matrix) <- make.unique(gene.labels)
  
  colnames(heatmap.matrix) <- gsub(
    "LFQ\\.intensity\\.",
    "",
    colnames(heatmap.matrix)
  )
  
  # Condition is displayed as a group heading rather than as a colour bar.
  annotation.col <- data.frame(
    Replicate = sub(".*_([ABC])_.*", "\\1", colnames(heatmap.matrix)),
    Measurement = sub(".*_([0-9]+)$", "\\1", colnames(heatmap.matrix))
  )
  
  rownames(annotation.col) <- colnames(heatmap.matrix)
  
  heatmap.matrix.z <- t(scale(t(heatmap.matrix)))
  heatmap.matrix.z <- heatmap.matrix.z[complete.cases(heatmap.matrix.z), , drop = FALSE]
  
  jpeg(
    file.path(output_dir, "Heatmap_significant_proteins_DreamAI_DEqMS.jpg"),
    width = 6000,
    height = 7000,
    res = 600,
    quality = 100
  )
  
  
  
  # Desired sample order
  desired.order <- c(
    "1G_A_01","1G_A_02","1G_A_03",
    "1G_B_01","1G_B_02","1G_B_03",
    "1G_C_01","1G_C_02","1G_C_03",
    "DLN_A_01","DLN_A_02","DLN_A_03",
    "DLN_B_01","DLN_B_02","DLN_B_03",
    "DLN_C_01","DLN_C_02","DLN_C_03"
  )
  
  # Reorder the significant-protein heatmap matrix
  available.order <- desired.order[desired.order %in% colnames(heatmap.matrix.z)]
  heatmap.matrix.z <- heatmap.matrix.z[, available.order, drop = FALSE]
  
  # Reorder annotations
  annotation.col <- annotation.col[available.order, , drop = FALSE]
  
  # Add a gap after each biological replicate. The central gap separates
  # MCM-1G from MCM-DLN.
  replicate.gaps <- c(3, 6, 9, 12, 15)
  replicate.gaps <- replicate.gaps[replicate.gaps < ncol(heatmap.matrix.z)]

  heatmap.plot <- pheatmap(
    heatmap.matrix.z,
    annotation_col = annotation.col,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    gaps_col = replicate.gaps,
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 7,
    fontsize_col = 8,
    main = NA,
    border_color = NA,
    silent = TRUE
  )

  # Place the condition names above their respective sample groups without
  # encoding condition as an annotation colour.
  matrix.layout <- heatmap.plot$gtable$layout[
    heatmap.plot$gtable$layout$name == "matrix",
  ]

  heatmap.plot$gtable <- gtable::gtable_add_rows(
    heatmap.plot$gtable,
    heights = grid::unit(0.45, "cm"),
    pos = matrix.layout$t - 1
  )

  heatmap.plot$gtable <- gtable::gtable_add_grob(
    heatmap.plot$gtable,
    grobs = grid::textGrob(
      label = c("MCM-1G", "MCM-DLN"),
      x = grid::unit(c(0.25, 0.75), "npc"),
      y = grid::unit(0.5, "npc"),
      gp = grid::gpar(fontface = "bold", fontsize = 10)
    ),
    t = matrix.layout$t,
    l = matrix.layout$l,
    r = matrix.layout$r,
    clip = "off",
    name = "condition_headings"
  )

  grid::grid.newpage()
  grid::grid.draw(heatmap.plot$gtable)
  
  dev.off()
}


if (exists("heatmap.matrix.z") && nrow(heatmap.matrix.z) > 1L) {
  write.table(
    heatmap.matrix.z,
    file.path(output_dir, "Heatmap_matrix_significant_proteins_zscore.txt"),
    sep = "\t", quote = FALSE
  )
}

# -----------------------------
# Individual protein abundance plot
# -----------------------------

gene_of_interest <- "STAT3"

hits <- rownames(df.prot)[
  grepl(gene_of_interest, df.prot$Gene.names)
]

if (length(hits) > 0) {
  
  protein_id <- hits[1]
  
  plot.df <- data.frame(
    Intensity = as.numeric(protein.matrix[protein_id, ]),
    Sample = colnames(protein.matrix),
    Condition = class
  )
  
  protein.plot <- ggplot(
    plot.df,
    aes(x = Condition, y = Intensity, color = Condition)
  ) +
    geom_boxplot(outlier.shape = NA, alpha = 0.2) +
    geom_jitter(width = 0.1, size = 3) +
    theme_bw(base_size = 14) +
    ggtitle(paste(gene_of_interest, protein_id)) +
    ylab("log2 LFQ intensity") +
    xlab("Condition")
  
  protein.plot
  
  ggsave(
    filename = file.path(output_dir, paste0(gene_of_interest, "_protein_abundance.jpg")),
    plot = protein.plot,
    width = 6,
    height = 5,
    dpi = 600
  )
  
  DEqMS.results[
    grepl(gene_of_interest, DEqMS.results$Gene.name),
    c("Gene.name", "logFC", "sca.P.Value", "sca.adj.pval")
  ]
  
} else {
  message(paste("Gene not found:", gene_of_interest))
}

```

## 11. Export summaries and combined figures

Export significant-protein table, dataset overview,
missingness comparison and multi-panel figures.

```{r summaries-and-panels, fig.width=12, fig.height=9}

sig.summary <- data.frame(
  Cutoff = paste0(
    "|log2FC| > ",
    logFC_cutoff,
    " and DEqMS FDR <= ",
    fdr_cutoff
  ),
  Significant_proteins = sum(
    DEqMS.results$sca.adj.pval <= fdr_cutoff &
      abs(DEqMS.results$logFC) > logFC_cutoff,
    na.rm = TRUE
  ),
  Up_in_MCM_1G = sum(
    DEqMS.results$sca.adj.pval <= fdr_cutoff &
      DEqMS.results$logFC > logFC_cutoff,
    na.rm = TRUE
  ),
  Up_in_MCM_DLN = sum(
    DEqMS.results$sca.adj.pval <= fdr_cutoff &
      DEqMS.results$logFC < -logFC_cutoff,
    na.rm = TRUE
  )
)

sig.summary

write.table(
  sig.summary,
  file.path(output_dir, "Significant_protein_summary_DreamAI_DEqMS.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# -----------------------------
# Dataset overview table
# -----------------------------

dataset.overview <- data.frame(
  Step = c("Raw MaxQuant", "After filtering", "After DreamAI"),
  Features = c(
    nrow(df.LFQ[, lfq_cols]),
    nrow(log2.lfq),
    nrow(protein.matrix)
  ),
  Samples = c(
    length(lfq_cols),
    ncol(log2.lfq),
    ncol(protein.matrix)
  )
)

dataset.overview

write.table(
  dataset.overview,
  file.path(output_dir, "Dataset_overview_features_samples.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# -----------------------------
# Missingness before vs after DreamAI
# -----------------------------

overall_missing_before <- mean(is.na(log2.lfq)) * 100
overall_missing_after <- mean(is.na(protein.matrix)) * 100

missing.before.after <- data.frame(
  Step = c("Before DreamAI", "After DreamAI"),
  Missing_values = c(
    sum(is.na(log2.lfq)),
    sum(is.na(protein.matrix))
  ),
  Missing_percent = round(
    c(
      overall_missing_before,
      overall_missing_after
    ),
    2
  )
)

missing.before.after

write.table(
  missing.before.after,
  file.path(output_dir, "Missingness_before_after_DreamAI.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# -----------------------------
# Multi-panel figures
# -----------------------------

qc.missing.plot <- missing.plot +
  ggtitle("A. Missing value distribution")

qc.density.plot <- density.plot +
  ggtitle("B. LFQ intensity density")

qc.pca.before.plot <- pca.before.plot +
  ggtitle("C. PCA before DreamAI")

qc.pca.after.plot <- pca.after.plot +
  ggtitle("D. PCA after DreamAI")

qc.volcano.plot <- volcano.plot +
  ggtitle("A. Volcano plot")

qc.ma.plot <- ma.plot +
  ggtitle("B. MA plot")

# Panel 1: missingness and imputation QC

jpeg(
  file.path(output_dir, "Panel_1_Missingness_Imputation_QC.jpg"),
  width = 9000,
  height = 7000,
  res = 600,
  quality = 100
)

grid.arrange(
  qc.missing.plot,
  qc.density.plot,
  qc.pca.before.plot,
  qc.pca.after.plot,
  ncol = 2,
  top = textGrob(
    "Panel 1: Missingness and DreamAI imputation QC",
    gp = gpar(fontsize = 18, fontface = "bold")
  )
)

dev.off()

# Panel 2: differential abundance

jpeg(
  file.path(output_dir, "Panel_2_Differential_Abundance.jpg"),
  width = 7000,
  height = 10000,
  res = 600,
  quality = 100
)

grid.arrange(
  qc.volcano.plot,
  qc.ma.plot,
  ncol = 1,
  heights = c(1, 1),
  top = textGrob(
    "Panel 2: Differential protein abundance analysis",
    gp = gpar(fontsize = 18, fontface = "bold")
  )
)

dev.off()

# Panel 3: DEqMS diagnostics

jpeg(
  file.path(output_dir, "Panel_3_DEqMS_Diagnostics.jpg"),
  width = 9000,
  height = 7000,
  res = 600,
  quality = 100
)

par(mfrow = c(2, 2), mar = c(5, 5, 4, 2))

VarianceScatterplot(
  fit4,
  xlab = "log2(Peptide count + 1)",
  ylab = "log(Variance)",
  main = "A. Variance scatterplot"
)

VarianceBoxplot(
  fit4,
  n = max(fit4$count, na.rm = TRUE),
  xlab = "Peptide count + 1",
  ylab = "log(Variance)",
  main = "B. Variance boxplot"
)

Residualplot(
  fit4,
  xlab = "log2(Peptide count + 1)",
  ylab = "Variance(fitted - observed)",
  main = "C. Residual plot"
)

plot.new()

par(mfrow = c(1, 1))

dev.off()

```

## 12. Record the computational environment

```{r session-information}

sessionInfo()
```

## 13. Exploratory sensitivity analyses

### 13.1 Prepare the exploratory workflow

```{r exploratory-setup}

# -----------------------------
# Exploratory output folders
# -----------------------------

exploratory_dir <- file.path(
  output_dir,
  "Exploratory_imputation_comparison"
)

impseq_dir <- file.path(
  exploratory_dir,
  "ImpSeq"
)

no_imputation_dir <- file.path(
  exploratory_dir,
  "No_imputation"
)

dir.create(
  impseq_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  no_imputation_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# -----------------------------
# Confirm required package
# -----------------------------

if (!requireNamespace("rrcovNA", quietly = TRUE)) {
  stop(
    paste0(
      "The rrcovNA package is required for ImpSeq. ",
      "Install it using install.packages('rrcovNA')."
    )
  )
}

# -----------------------------
# Reusable exploratory DEqMS function
# -----------------------------

run_exploratory_deqms <- function(
    expression_matrix,
    analysis_name,
    analysis_title,
    analysis_output_dir,
    df_prot,
    peptide_count_table,
    logFC_cutoff = 1.0,
    fdr_cutoff = 0.05
) {
  
  # Ensure matrix format
  expression_matrix <- as.matrix(expression_matrix)
  storage.mode(expression_matrix) <- "numeric"
  
  # Remove proteins containing non-finite values
  valid_rows <- apply(
    expression_matrix,
    1,
    function(x) all(is.finite(x))
  )
  
  expression_matrix <- expression_matrix[
    valid_rows,
    ,
    drop = FALSE
  ]
  
  if (nrow(expression_matrix) < 2) {
    stop(
      "Too few proteins available for ",
      analysis_name,
      " analysis."
    )
  }
  
  # ----------------------------------
  # Construct group factor and design
  # ----------------------------------
  
  exploratory_class <- ifelse(
    grepl(
      "^LFQ\\.intensity\\.1G_",
      colnames(expression_matrix)
    ),
    "MCM_1G",
    "MCM_DLN"
  )
  
  exploratory_class <- factor(
    exploratory_class,
    levels = c(
      "MCM_1G",
      "MCM_DLN"
    )
  )
  
  exploratory_design <- model.matrix(
    ~ 0 + exploratory_class
  )
  
  colnames(exploratory_design) <- levels(
    exploratory_class
  )
  
  # ----------------------------------
  # Fit limma model
  # ----------------------------------
  
  exploratory_fit1 <- lmFit(
    expression_matrix,
    design = exploratory_design
  )
  
  exploratory_contrast <- makeContrasts(
    MCM_1G - MCM_DLN,
    levels = exploratory_design
  )
  
  exploratory_fit2 <- contrasts.fit(
    exploratory_fit1,
    contrasts = exploratory_contrast
  )
  
  exploratory_fit3 <- eBayes(
    exploratory_fit2
  )
  
  # ----------------------------------
  # Add peptide counts for DEqMS
  # ----------------------------------
  
  exploratory_fit3$count <- peptide_count_table[
    rownames(exploratory_fit3$coefficients),
    "count"
  ]
  
  if (any(is.na(exploratory_fit3$count))) {
    
    missing_count_ids <- rownames(
      exploratory_fit3$coefficients
    )[is.na(exploratory_fit3$count)]
    
    stop(
      paste0(
        "Missing peptide counts for ",
        length(missing_count_ids),
        " proteins in ",
        analysis_name,
        "."
      )
    )
  }
  
  if (any(exploratory_fit3$count <= 0)) {
    stop(
      "Non-positive peptide counts detected in ",
      analysis_name,
      "."
    )
  }
  
  # ----------------------------------
  # Fit DEqMS model
  # ----------------------------------
  
  exploratory_fit4 <- spectraCounteBayes(
    exploratory_fit3,
    fit.method = "loess",
    coef_col = 1
  )
  
  exploratory_results <- outputResult(
    exploratory_fit4,
    coef_col = 1
  )
  
  # ----------------------------------
  # Add gene names
  # ----------------------------------
  
  exploratory_results$Gene.name <- df_prot[
    exploratory_results$gene,
    "Gene.names"
  ]
  
  missing_gene_names <- (
    is.na(exploratory_results$Gene.name) |
      exploratory_results$Gene.name == ""
  )
  
  exploratory_results$Gene.name[
    missing_gene_names
  ] <- exploratory_results$gene[
    missing_gene_names
  ]
  
  # ----------------------------------
  # Classify proteins
  # ----------------------------------
  
  exploratory_results$significant <- with(
    exploratory_results,
    sca.adj.pval <= fdr_cutoff &
      abs(logFC) > logFC_cutoff
  )
  
  exploratory_results$Regulation <- "Not significant"
  
  exploratory_results$Regulation[
    exploratory_results$logFC > logFC_cutoff &
      exploratory_results$sca.adj.pval <= fdr_cutoff
  ] <- "Up in MCM-1G"
  
  exploratory_results$Regulation[
    exploratory_results$logFC < -logFC_cutoff &
      exploratory_results$sca.adj.pval <= fdr_cutoff
  ] <- "Up in MCM-DLN"
  
  exploratory_results$Regulation <- factor(
    exploratory_results$Regulation,
    levels = c(
      "Up in MCM-DLN",
      "Up in MCM-1G",
      "Not significant"
    )
  )
  
  # Avoid infinite values if adjusted p-value is zero
  exploratory_results$plot.adj.pval <- pmax(
    exploratory_results$sca.adj.pval,
    .Machine$double.xmin
  )
  
  exploratory_results$minus.log10.adj.pval <- -log10(
    exploratory_results$plot.adj.pval
  )
  
  # ----------------------------------
  # Save complete DEqMS result table
  # ----------------------------------
  
  write.table(
    exploratory_results,
    file.path(
      analysis_output_dir,
      paste0(
        "DEqMS_results_",
        analysis_name,
        ".txt"
      )
    ),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
  # ----------------------------------
  # Save significant proteins
  # ----------------------------------
  
  exploratory_significant <- exploratory_results[
    exploratory_results$significant,
    ,
    drop = FALSE
  ]
  
  exploratory_significant <- exploratory_significant[
    order(
      exploratory_significant$sca.adj.pval,
      -abs(exploratory_significant$logFC)
    ),
    ,
    drop = FALSE
  ]
  
  write.table(
    exploratory_significant,
    file.path(
      analysis_output_dir,
      paste0(
        "Significant_proteins_",
        analysis_name,
        ".txt"
      )
    ),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
  # ----------------------------------
  # Select labels for volcano plot
  # ----------------------------------
  
  exploratory_labels <- exploratory_results[
    exploratory_results$significant &
      !is.na(exploratory_results$Gene.name) &
      exploratory_results$Gene.name != "" &
      is.finite(
        exploratory_results$minus.log10.adj.pval
      ),
    ,
    drop = FALSE
  ]
  
  exploratory_labels <- exploratory_labels[
    order(exploratory_labels$sca.adj.pval),
    ,
    drop = FALSE
  ]
  
  exploratory_labels <- head(
    exploratory_labels,
    20
  )
  
  # ----------------------------------
  # Volcano plot
  # ----------------------------------
  
  exploratory_volcano <- ggplot(
    exploratory_results,
    aes(
      x = logFC,
      y = minus.log10.adj.pval
    )
  ) +
    geom_point(
      aes(color = Regulation),
      size = 0.6
    ) +
    geom_vline(
      xintercept = c(
        -logFC_cutoff,
        logFC_cutoff
      ),
      colour = "red"
    ) +
    geom_hline(
      yintercept = -log10(fdr_cutoff),
      colour = "red"
    ) +
    geom_vline(
      xintercept = 0,
      colour = "black"
    ) +
    scale_colour_manual(
      values = c(
        "Up in MCM-DLN" = "blue",
        "Up in MCM-1G" = "orange",
        "Not significant" = "grey75"
      )
    ) +
    theme_bw(base_size = 16) +
    labs(
      title = analysis_title,
      x = expression(
        log[2] * "(MCM-1G / MCM-DLN)"
      ),
      y = expression(
        -log[10] * "(DEqMS adjusted p-value)"
      ),
      colour = "Protein regulation",
      caption = paste0(
        "Exploratory analysis; cutoffs: |log2FC| > ",
        logFC_cutoff,
        " and DEqMS adjusted p-value <= ",
        fdr_cutoff
      )
    )
  
  if (nrow(exploratory_labels) > 0) {
    
    exploratory_volcano <- exploratory_volcano +
      geom_text_repel(
        data = exploratory_labels,
        aes(label = Gene.name),
        max.overlaps = 20,
        size = 3
      )
  }
  
  ggsave(
    filename = file.path(
      analysis_output_dir,
      paste0(
        "Volcano_plot_",
        analysis_name,
        "_DEqMS.jpg"
      )
    ),
    plot = exploratory_volcano,
    width = 8,
    height = 7,
    dpi = 600
  )
  
  # ----------------------------------
  # Analysis summary
  # ----------------------------------
  
  exploratory_summary <- data.frame(
    Analysis = analysis_name,
    Proteins_tested = nrow(exploratory_results),
    Significant_proteins = sum(
      exploratory_results$significant,
      na.rm = TRUE
    ),
    Up_in_MCM_1G = sum(
      exploratory_results$Regulation == "Up in MCM-1G",
      na.rm = TRUE
    ),
    Up_in_MCM_DLN = sum(
      exploratory_results$Regulation == "Up in MCM-DLN",
      na.rm = TRUE
    ),
    Log2FC_cutoff = logFC_cutoff,
    FDR_cutoff = fdr_cutoff,
    stringsAsFactors = FALSE
  )
  
  write.table(
    exploratory_summary,
    file.path(
      analysis_output_dir,
      paste0(
        "Summary_",
        analysis_name,
        ".txt"
      )
    ),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
  return(
    list(
      results = exploratory_results,
      significant = exploratory_significant,
      summary = exploratory_summary,
      volcano = exploratory_volcano
    )
  )
}

```

### 13.2 ImpSeq analysis

ImpSeq is applied to the same filtered log2 LFQ matrix.

```{r impseq-sensitivity}

message(
  "Starting exploratory ImpSeq imputation and DEqMS analysis."
)

# impSeq operates on the filtered log2 LFQ matrix.
# Proteins are rows and measurement replicates are columns.

impseq.matrix <- rrcovNA::impSeq(
  as.matrix(log2.lfq)
)

impseq.matrix <- as.matrix(
  impseq.matrix
)

storage.mode(impseq.matrix) <- "numeric"

rownames(impseq.matrix) <- rownames(
  log2.lfq
)

colnames(impseq.matrix) <- colnames(
  log2.lfq
)

# Confirm that imputation produced a complete matrix
if (anyNA(impseq.matrix)) {
  stop(
    "ImpSeq did not replace all missing values."
  )
}

if (any(!is.finite(impseq.matrix))) {
  stop(
    "ImpSeq produced non-finite values."
  )
}

# Save ImpSeq-imputed matrix
write.table(
  impseq.matrix,
  file.path(
    impseq_dir,
    "ImpSeq_imputed_Melanoma_LFQ.tsv"
  ),
  sep = "\t",
  quote = FALSE
)

# Run only DEqMS and volcano plot
impseq.analysis <- run_exploratory_deqms(
  expression_matrix = impseq.matrix,
  analysis_name = "ImpSeq",
  analysis_title = paste0(
    "Exploratory DEqMS analysis after ImpSeq imputation"
  ),
  analysis_output_dir = impseq_dir,
  df_prot = df.prot,
  peptide_count_table = pep.count.table,
  logFC_cutoff = logFC_cutoff,
  fdr_cutoff = fdr_cutoff
)

```

### 13.3 Complete-case sensitivity analysis

limma and DEqMS require complete samples - restrict

```{r complete-case-sensitivity}

message(
  paste0(
    "Starting exploratory DEqMS analysis without imputation. ",
    "Only proteins complete across all samples are included."
  )
)

# limma and DEqMS require a complete expression matrix.
# Therefore, the no-imputation analysis is restricted to
# proteins without missing LFQ values across all 18 samples.

no.imputation.matrix <- log2.lfq[
  complete.cases(log2.lfq),
  ,
  drop = FALSE
]

if (nrow(no.imputation.matrix) < 2) {
  stop(
    paste0(
      "Fewer than two proteins have complete LFQ measurements ",
      "across all samples. The no-imputation analysis cannot ",
      "be performed."
    )
  )
}

message(
  "Proteins retained without imputation: ",
  nrow(no.imputation.matrix)
)

# Save complete-case matrix
write.table(
  no.imputation.matrix,
  file.path(
    no_imputation_dir,
    "Complete_case_Melanoma_LFQ_no_imputation.tsv"
  ),
  sep = "\t",
  quote = FALSE
)

# Run only DEqMS and volcano plot
no.imputation.analysis <- run_exploratory_deqms(
  expression_matrix = no.imputation.matrix,
  analysis_name = "No_imputation_complete_cases",
  analysis_title = paste0(
    "Exploratory DEqMS analysis without imputation"
  ),
  analysis_output_dir = no_imputation_dir,
  df_prot = df.prot,
  peptide_count_table = pep.count.table,
  logFC_cutoff = logFC_cutoff,
  fdr_cutoff = fdr_cutoff
)

```

### 13.4 Compare sensitivity-analysis results


```{r sensitivity-summary}

exploratory.comparison <- rbind(
  data.frame(
    Method = "DreamAI",
    Proteins_tested = nrow(DEqMS.results),
    Significant_proteins = sum(
      DEqMS.results$sca.adj.pval <= fdr_cutoff &
        abs(DEqMS.results$logFC) > logFC_cutoff,
      na.rm = TRUE
    ),
    Up_in_MCM_1G = sum(
      DEqMS.results$sca.adj.pval <= fdr_cutoff &
        DEqMS.results$logFC > logFC_cutoff,
      na.rm = TRUE
    ),
    Up_in_MCM_DLN = sum(
      DEqMS.results$sca.adj.pval <= fdr_cutoff &
        DEqMS.results$logFC < -logFC_cutoff,
      na.rm = TRUE
    )
  ),
  data.frame(
    Method = "ImpSeq",
    Proteins_tested =
      impseq.analysis$summary$Proteins_tested,
    Significant_proteins =
      impseq.analysis$summary$Significant_proteins,
    Up_in_MCM_1G =
      impseq.analysis$summary$Up_in_MCM_1G,
    Up_in_MCM_DLN =
      impseq.analysis$summary$Up_in_MCM_DLN
  ),
  data.frame(
    Method = "No imputation (complete cases)",
    Proteins_tested =
      no.imputation.analysis$summary$Proteins_tested,
    Significant_proteins =
      no.imputation.analysis$summary$Significant_proteins,
    Up_in_MCM_1G =
      no.imputation.analysis$summary$Up_in_MCM_1G,
    Up_in_MCM_DLN =
      no.imputation.analysis$summary$Up_in_MCM_DLN
  )
)

write.table(
  exploratory.comparison,
  file.path(
    exploratory_dir,
    "DreamAI_ImpSeq_NoImputation_comparison.txt"
  ),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

print(exploratory.comparison)

message(
  paste0(
    "Exploratory ImpSeq and no-imputation analyses completed. ",
    "No PCA, MA plot, heatmap, pathway analysis, or other ",
    "downstream analysis was performed for these alternatives."
  )
)
```
