---
title: "Melanoma DIABLO Study"
params:
  sers_selection_method: "SGL"
  ws_correlation_threshold: 0.99
output:
  pdf_document: default
  html_document:
    self_contained: true
    df_print: paged
---

# *DIABLO Melanoma Multi-Omics Study*
This workflow integrates RNA-seq, LFQ proteomics and SERS data using DIABLO. The aim is to identify a correlated multi-omics signature that separates the two melanoma cell lines `1G` and `DLN`.
The transcriptomics part uses only the 2D cell samples. The proteomics part uses the MaxQuant `proteinGroups` file where technical replicates were already merged.
In total six groups will be retained across the three omics layers: 1G_A/_B/_C and DLN_A/_B/_C.

Note: This script provides three approaches for reducing the SERS feature space: sparse group lasso (`SGL`), a genetic algorithm (`GA`), and correlation-based variable window selection (`WS`). The method is selected with the `sers_selection_method` parameter in the YAML header.

## Setup
Standardize plot parameters and define the project root directory.

```{r setup, include=FALSE}
setwd("../DIABLO_Melanoma_Pauser")

library(knitr)

opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  dev = "png",
  fig.retina = 2,
  fig.path = "figs-html/",
  fig.keep = "all",
  out.width = "100%",
  fig.align = "center"
)

rmd_input <- knitr::current_input()

if (!is.null(rmd_input) && !is.na(rmd_input)) {
  knitr::opts_knit$set(
    root.dir = dirname(normalizePath(rmd_input))
  )
}

set.seed(123)
```

## Load libraries

```{r packages}
library(mixOmics)
library(DESeq2)
library(edgeR)
library(dplyr)
library(igraph)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(BiocParallel)
library(msgl)
library(GA)
library(parallel)
library(doParallel)

if (!requireNamespace("ComplexHeatmap", quietly = TRUE)) {
  stop(
    "Package 'ComplexHeatmap' is required for the labelled DIABLO heatmap. ",
    "Install it with BiocManager::install('ComplexHeatmap')."
  )
}

if (!requireNamespace("circlize", quietly = TRUE)) {
  stop(
    "Package 'circlize' is required for the labelled DIABLO heatmap. ",
    "Install it with install.packages('circlize')."
  )
}

set.seed(123)

sers_selection_method <- toupper(params$sers_selection_method)
ws_correlation_threshold <- as.numeric(params$ws_correlation_threshold)

if (!sers_selection_method %in% c("SGL", "GA", "WS")) {
  stop("sers_selection_method must be 'SGL', 'GA', or 'WS'.")
}

if (
  length(ws_correlation_threshold) != 1L ||
  !is.finite(ws_correlation_threshold) ||
  ws_correlation_threshold < -1 ||
  ws_correlation_threshold > 1
) {
  stop("ws_correlation_threshold must be a number between -1 and 1.")
}

cat("SERS region-selection method:", sers_selection_method, "\n")

if (sers_selection_method == "WS") {
  cat("WS correlation threshold:", ws_correlation_threshold, "\n")
}
```

## Define input directories and file paths

```{r paths}
project_root <- getwd()

rna_dir  <- file.path(project_root, "Transcriptome")
prot_dir <- file.path(project_root, "Proteome")
sers_dir <- file.path(project_root, "SERS")

sers_method_tag <- toupper(sers_selection_method)

results_dir <- file.path(
  project_root,
  paste0("DIABLO_", sers_method_tag, "_results")
)

figure_dir <- file.path(
  results_dir,
  paste0(sers_method_tag, "_DIABLO_figures")
)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

result_file <- function(filename) {
  file.path(
    results_dir,
    paste0(sers_method_tag, "_", filename)
  )
}

cat("Method-specific results directory:", results_dir, "\n")

prot_file <- file.path(
  prot_dir,
  "proteinGroups_results_run1_mqpar_2025__FH_250515_Mayr_extracts_mcm_new_merged.txt.txt"
)

sers_file <- file.path(
  sers_dir,
  "melamonma_g_dln_sers_rerun_annamasterthesis_preprocessed.csv"
)

stopifnot(
  dir.exists(rna_dir),
  dir.exists(prot_dir),
  dir.exists(sers_dir),
  file.exists(prot_file),
  file.exists(sers_file)
)
```


## Helper functions for sample-name harmonisation and data aggregation

```{r helper-functions}
clean_sample_names <- function(x) {
  x <- basename(x)
  x <- gsub("[.]bam$|[.]sorted$|[.]rmdup$|[.]counts$|[.]txt$|[.]tsv$|[.]csv$", "", x, ignore.case = TRUE)
  x <- gsub("[.]featurecount$|[.]featurecounts$|_featurecount$|_featurecounts$|_counts$", "", x, ignore.case = TRUE)
  x <- gsub("LFQ[.]intensity[.]", "", x, ignore.case = TRUE)
  x <- gsub("Intensity[.]", "", x, ignore.case = TRUE)
  x <- gsub("Reporter[.]intensity[.]", "", x, ignore.case = TRUE)
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

make_unique <- function(x) {
  x <- ifelse(is.na(x) | x == "", "feature", x)
  make.unique(x)
}

get_group <- function(x) {
  sapply(strsplit(as.character(x), "_"), function(z) z[1])
}

get_biorep <- function(x) {
  sapply(strsplit(as.character(x), "_"), function(z) ifelse(length(z) >= 2, z[2], NA_character_))
}

collapse_columns_by_name_sum <- function(mat) {
  ids <- colnames(mat)
  ids_unique <- unique(ids)
  
  collapsed <- sapply(ids_unique, function(id) {
    cols <- which(ids == id)
    rowSums(mat[, cols, drop = FALSE], na.rm = TRUE)
  })
  
  collapsed <- as.matrix(collapsed)
  colnames(collapsed) <- ids_unique
  collapsed
}

harmonise_rna_sample_names <- function(x) {
  x <- basename(x)
  x <- gsub("[.]featurecount$|[.]featurecounts$", "", x, ignore.case = TRUE)
  x <- gsub("_STAR.*$", "", x, ignore.case = TRUE)
  x <- gsub("-", "_", x)
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  
  rep_map <- c("1" = "A", "2" = "B", "3" = "C")
  out <- rep(NA_character_, length(x))
  
  for (i in seq_along(x)) {
    parts <- strsplit(x[i], "_")[[1]]
    
    if (length(parts) == 2 && parts[1] == "1G" && parts[2] %in% names(rep_map)) {
      out[i] <- paste0("1G_", rep_map[parts[2]])
    }
    
    if (length(parts) == 3 && parts[1] == "DLN" && parts[2] == "2D" && parts[3] %in% names(rep_map)) {
      out[i] <- paste0("DLN_", rep_map[parts[3]])
    }
  }
  
  out
}

harmonise_sers_sample_names <- function(x) {
  x <- basename(x)
  group <- ifelse(grepl("_1G_", x), "1G", ifelse(grepl("_DLN_", x), "DLN", NA_character_))
  m <- sub(".*_M([123])[.].*$", "\\1", x)
  rep_map <- c("1" = "A", "2" = "B", "3" = "C")
  rep <- rep_map[m]
  ifelse(!is.na(group) & !is.na(rep), paste0(group, "_", rep), NA_character_)
}

harmonise_prot_sample_names <- function(x) {
  x <- clean_sample_names(x)
  x <- gsub("^LFQ_intensity_", "", x, ignore.case = TRUE)
  x <- gsub("^LFQ_intensity", "", x, ignore.case = TRUE)
  x <- gsub("^Intensity_", "", x, ignore.case = TRUE)
  x <- gsub("^Reporter_intensity_", "", x, ignore.case = TRUE)
  x <- gsub("_2D_", "_", x, ignore.case = TRUE)
  x <- gsub("DLN_2D_", "DLN_", x, ignore.case = TRUE)
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  
  rep_map <- c("1" = "A", "2" = "B", "3" = "C")
  out <- rep(NA_character_, length(x))
  
  for (i in seq_along(x)) {
    parts <- strsplit(x[i], "_")[[1]]
    
    if (length(parts) >= 2 && parts[1] == "1G") {
      rep_token <- parts[2]
      if (rep_token %in% names(rep_map)) rep_token <- rep_map[rep_token]
      if (rep_token %in% c("A", "B", "C")) out[i] <- paste0("1G_", rep_token)
    }
    
    if (length(parts) >= 2 && parts[1] == "DLN") {
      rep_token <- parts[length(parts)]
      if (rep_token %in% names(rep_map)) rep_token <- rep_map[rep_token]
      if (rep_token %in% c("A", "B", "C")) out[i] <- paste0("DLN_", rep_token)
    }
  }
  out
}
```

## Define and verify the RNA-seq input files

```{r rna-load}
rna_files <- file.path(
  rna_dir,
  c(
    "1G-1_STAR_WGF_sorted.rmdup.featurecount",
    "1G-2_STAR_WGF_sorted.rmdup.featurecount",
    "1G-3_STAR_WGF_sorted.rmdup.featurecount",
    "DLN-2D-1_STAR_WGF_sorted.rmdup.featurecount",
    "DLN-2D-2_STAR_WGF_sorted.rmdup.featurecount",
    "DLN-2D-3_STAR_WGF_sorted.rmdup.featurecount"
  )
)

missing_rna_files <- rna_files[!file.exists(rna_files)]

if (length(missing_rna_files) > 0) {
  stop(
    "Missing RNA-seq files: ",
    paste(missing_rna_files, collapse = ", ")
  )
}

stopifnot(length(rna_files) == 6)
```

## Read, merge, and format the RNA-seq count data

```{r rna-read}
read_featurecounts <- function(f) {
  x <- read.delim(f, comment.char = "#", check.names = FALSE)
  if (!"Geneid" %in% colnames(x)) stop(paste("No Geneid column in", f))
  
  annotation_cols <- intersect(c("Geneid", "Chr", "Start", "End", "Strand", "Length"), colnames(x))
  count_cols <- setdiff(colnames(x), annotation_cols)
  count_cols <- count_cols[sapply(x[count_cols], is.numeric)]
  
  out <- x[, c("Geneid", count_cols), drop = FALSE]
  colnames(out)[-1] = harmonise_rna_sample_names(f)
  out
}

rna_list <- lapply(rna_files, read_featurecounts)

if (length(rna_list) == 1) {
  rna_counts <- rna_list[[1]]
} else {
  rna_counts <- Reduce(function(a, b) full_join(a, b, by = "Geneid"), rna_list)
}

rna_counts <- as.data.frame(rna_counts)
rna_gene <- rna_counts$Geneid
rna_counts <- rna_counts[, -1, drop = FALSE]
rownames(rna_counts) = make_unique(rna_gene)

rna_counts[is.na(rna_counts)] = 0
rna_counts <- round(as.matrix(rna_counts))
storage.mode(rna_counts) = "integer"

rna_counts <- collapse_columns_by_name_sum(rna_counts)
storage.mode(rna_counts) = "integer"

expected_samples <- c("1G_A", "1G_B", "1G_C", "DLN_A", "DLN_B", "DLN_C")
rna_counts <- rna_counts[, expected_samples, drop = FALSE]

cat("RNA samples retained:\n")
print(colnames(rna_counts))

dim(rna_counts)
rna_counts[1:5, 1:min(5, ncol(rna_counts))]
```

## Load and inspect the proteomics data.

```{r prot-load}
if (!file.exists(prot_file)) {
  stop(paste("Proteomics file does not exist:", prot_file))
}

prot_raw <- read.delim(prot_file, check.names = FALSE)
dim(prot_raw)
head(prot_raw[, 1:min(6, ncol(prot_raw))])
```

# Format, label, and filter the proteomics data without imputation.

```{r prot-format}
feature_cols <- grep(
  "gene|protein|uniprot|accession|id|name|description",
  colnames(prot_raw),
  ignore.case = TRUE,
  value = TRUE
)

numeric_cols <- colnames(prot_raw)[
  vapply(prot_raw, is.numeric, logical(1))
]

preferred_feature_cols <- c(
  "Gene names", "Gene.names", "Gene name", "Gene.name",
  "Protein names", "Protein.names",
  "Majority protein IDs", "Majority.protein.IDs",
  "Protein IDs", "Protein.IDs"
)

available_feature_cols <- preferred_feature_cols[
  preferred_feature_cols %in% colnames(prot_raw)
]

if (length(available_feature_cols) > 0) {
  feature_col <- available_feature_cols[1]
} else if (length(feature_cols) > 0) {
  feature_col <- feature_cols[1]
} else {
  feature_col <- colnames(prot_raw)[1]
}

# Identify and harmonize sample columns
sample_cols <- grep(
  "LFQ[.]intensity|LFQ intensity|Intensity|Reporter[.]intensity",
  numeric_cols,
  ignore.case = TRUE,
  value = TRUE
)

if (length(sample_cols) == 0) {
  sample_cols <- numeric_cols
}

sample_names <- harmonise_prot_sample_names(sample_cols)
keep_samples <- !is.na(sample_names)

sample_cols <- sample_cols[keep_samples]
sample_names <- sample_names[keep_samples]

stopifnot(length(sample_cols) >= 6)

# Create the protein-abundance matrix
prot_mat <- as.matrix(prot_raw[, sample_cols, drop = FALSE])
storage.mode(prot_mat) <- "double"

protein_labels <- as.character(prot_raw[[feature_col]])
missing_labels <- is.na(protein_labels) | protein_labels == ""
protein_labels[missing_labels] <- as.character(prot_raw[[1]])[missing_labels]

protein_labels <- gsub(";.*$", "", protein_labels)
protein_labels <- gsub("\\s+", "_", protein_labels)
protein_labels <- substr(protein_labels, 1, 40)

rownames(prot_mat) <- make_unique(protein_labels)
colnames(prot_mat) <- sample_names

# Retain and order the six biological samples
prot_mat <- prot_mat[
  ,
  expected_samples,
  drop = FALSE
]

# MaxQuant represents missing LFQ intensities as either zero or NA
prot_mat[prot_mat == 0] <- NA_real_

# Exclude proteins with missing values; no imputation is applied
prot_mat <- prot_mat[
  complete.cases(prot_mat),
  ,
  drop = FALSE
]

# Remove proteins without variation across samples
prot_mat <- prot_mat[
  apply(prot_mat, 1, sd) > 0,
  ,
  drop = FALSE
]

cat("Protein-label column:", feature_col, "\n")
cat("Retained samples:\n")
print(colnames(prot_mat))

dim(prot_mat)
prot_mat[1:5, 1:min(5, ncol(prot_mat))]
```

## Load and format the SERS data.

```{r sers-import}
if (!file.exists(sers_file)) {
  stop(paste("SERS file does not exist:", sers_file))
}

sers_raw <- read.csv(
  sers_file,
  check.names = FALSE
)

stopifnot(all(c("label", "file") %in% colnames(sers_raw)))

sers_feature_cols <- setdiff(
  colnames(sers_raw),
  c("label", "file")
)

sers_feature_cols <- sers_feature_cols[
  vapply(sers_raw[sers_feature_cols], is.numeric, logical(1))
]

if (length(sers_feature_cols) < 2) {
  stop("Fewer than two numeric SERS variables were detected.")
}

sers_sample_id <- harmonise_sers_sample_names(sers_raw$file)
keep_sers <- !is.na(sers_sample_id)

sers_x_raw <- as.matrix(
  sers_raw[keep_sers, sers_feature_cols, drop = FALSE]
)

storage.mode(sers_x_raw) <- "double"
sers_sample_id <- sers_sample_id[keep_sers]

cat("Spectra per biological sample:\n")
print(table(sers_sample_id))

sers_classes <- factor(
  ifelse(grepl("^1G_", sers_sample_id), "1G", "DLN"),
  levels = c("1G", "DLN")
)

stopifnot(
  nrow(sers_x_raw) == length(sers_classes),
  ncol(sers_x_raw) == length(sers_feature_cols),
  !anyNA(sers_x_raw),
  all(is.finite(sers_x_raw)),
  length(levels(droplevels(sers_classes))) == 2
)

cat("\nSERS class distribution:\n")
print(table(sers_classes))

sers_wavenumbers_raw <- suppressWarnings(
  as.numeric(
    gsub(
      "SERS_X|SERS_|X|cm",
      "",
      colnames(sers_x_raw)
    )
  )
)

if (anyNA(sers_wavenumbers_raw)) {
  invalid_names <- colnames(sers_x_raw)[is.na(sers_wavenumbers_raw)]

  stop(
    paste(
      "Could not extract numeric Raman shifts from:",
      paste(invalid_names, collapse = ", ")
    )
  )
}

sers_order <- order(sers_wavenumbers_raw)

sers_wavenumbers_raw <- sers_wavenumbers_raw[sers_order]
sers_x_raw <- sers_x_raw[, sers_order, drop = FALSE]

colnames(sers_x_raw) <- make.unique(
  paste0(
    "SERS_",
    format(
      sers_wavenumbers_raw,
      trim = TRUE,
      scientific = FALSE
    )
  )
)

# Ten-wavenumber groups preserve local spectral structure for SGL and GA.
sgl_window_width <- 10

sers_grouping <- floor(
  (
    sers_wavenumbers_raw -
      min(sers_wavenumbers_raw)
  ) / sgl_window_width
) + 1L

stopifnot(
  length(sers_grouping) == ncol(sers_x_raw),
  all(sers_grouping >= 1)
)

cat("\nSERS input dimensions:\n")
print(dim(sers_x_raw))

cat("Number of fixed-width spectral groups:\n")
print(length(unique(sers_grouping)))

cat("Variables per spectral group:\n")
print(summary(as.numeric(table(sers_grouping))))

```

##Define replicate-aware cross-validation folds.

```{r sers-cv}
sers_biorep <- sub(
  "^.*_",
  "",
  sers_sample_id
)

stopifnot(all(sers_biorep %in% c("A", "B", "C")))

# Hold out the corresponding 1G and DLN replicate together.
sgl_cv_indices <- lapply(
  c("A", "B", "C"),
  function(rep_id) {
    which(sers_biorep == rep_id)
  }
)

names(sgl_cv_indices) <- paste0(
  "Leave_out_",
  c("A", "B", "C")
)

cat("\nSGL cross-validation fold sizes:\n")
print(vapply(sgl_cv_indices, length, integer(1)))

cat("\nClass distribution within validation folds:\n")
print(
  lapply(
    sgl_cv_indices,
    function(idx) table(sers_classes[idx])
  )
)

# Ensure that both classes occur in every validation fold
fold_has_both_classes <- vapply(
  sgl_cv_indices,
  function(idx) {
    length(unique(sers_classes[idx])) == 2
  },
  logical(1)
)

if (!all(fold_has_both_classes)) {
  stop("At least one validation fold does not contain both classes.")
}

```
## Fit and evaluate the Sparse Group LASSO models.

```{r sers-sgl}
set.seed(123)

if (sers_selection_method == "SGL") {

detected_cores <- parallel::detectCores(logical = FALSE)

if (is.na(detected_cores)) {
  detected_cores <- 1
}

n_cores <- max(
  1,
  min(2, detected_cores)
)

cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)

sgl_cv <- tryCatch(
  {
    msgl::cv(
      x = sers_x_raw,
      classes = sers_classes,
      grouping = sers_grouping,

      # alpha = 0: Group LASSO
      # alpha = 1: conventional LASSO
      # 0 < alpha < 1: Sparse Group LASSO
      alpha = 0.5,

      # Fit 100 models from lambda.max to
      # 0.1 * lambda.max
      lambda = 0.1,
      d = 100,

      fold = 3,
      cv.indices = sgl_cv_indices,

      standardize = TRUE,
      use_parallel = TRUE
    )
  },
  finally = {
    parallel::stopCluster(cl)
    foreach::registerDoSEQ()
  }
)

print(sgl_cv)

# Determine the best cross-validated model

sgl_error <- sglOptim::Err(sgl_cv)

if (
  length(sgl_error) == 0 ||
  all(!is.finite(sgl_error))
) {
  stop("No valid cross-validation errors were returned by msgl.")
}

minimum_cv_error <- min(
  sgl_error,
  na.rm = TRUE
)

# Prefer the sparsest model when several candidates have the same CV error.
best_candidates <- which(
  is.finite(sgl_error) &
    abs(sgl_error - minimum_cv_error) < sqrt(.Machine$double.eps)
)

cat("Tied CV models:", paste(best_candidates, collapse = ", "), "\n")

cat(
  "Minimum CV error:",
  minimum_cv_error,
  "\n"
)

}

```

##Extract the selected spectral features and regions

```{r sers-selection}
set.seed(123)

if (sers_selection_method == "SGL") {

sgl_fit <- msgl::fit(
  x = sers_x_raw,
  classes = sers_classes,
  grouping = sers_grouping,
  alpha = 0.5,
  lambda = 0.1,
  d = 100,
  standardize = TRUE
)

sgl_feature_path <- features(sgl_fit)

if (length(sgl_feature_path) != length(sgl_error)) {
  stop(
    paste(
      "The fitted model path and CV error path have different lengths:",
      length(sgl_feature_path),
      "versus",
      length(sgl_error)
    )
  )
}

# Exclude the intercept when comparing tied models.
candidate_feature_counts <- vapply(
  best_candidates,
  function(model_index) {

    selected <- setdiff(
      sgl_feature_path[[model_index]],
      "Intercept"
    )

    sum(selected %in% colnames(sers_x_raw))
  },
  integer(1)
)

candidate_summary <- data.frame(
  model_index = best_candidates,
  cv_error = sgl_error[best_candidates],
  selected_wavenumbers = candidate_feature_counts,
  stringsAsFactors = FALSE
)

cat("Tied-model summary:\n")
print(candidate_summary)

sgl_best_index <- best_candidates[
  which.min(candidate_feature_counts)
]

cat(
  "Selected SGL model:",
  sgl_best_index,
  "\n"
)

cat(
  "CV error:",
  sgl_error[sgl_best_index],
  "\n"
)

sers_selected_features <- setdiff(
  sgl_feature_path[[sgl_best_index]],
  "Intercept"
)

sers_selected_features <- intersect(
  sers_selected_features,
  colnames(sers_x_raw)
)

sers_active_grouping <- sers_grouping

if (length(sers_selected_features) == 0) {
  stop(
    paste(
      "The selected SGL model contains no SERS variables.",
      "Consider reducing lambda, for example from 0.1 to 0.05."
    )
  )
}

cat(
  "Selected SERS wavenumbers:",
  length(sers_selected_features),
  "\n"
)

print(sers_selected_features)

selected_idx <- match(
  sers_selected_features,
  colnames(sers_x_raw)
)

stopifnot(!anyNA(selected_idx))

sers_sgl_selection <- data.frame(
  feature = sers_selected_features,
  wavenumber = sers_wavenumbers_raw[selected_idx],
  group = sers_grouping[selected_idx],
  stringsAsFactors = FALSE
)

sers_sgl_selection <- sers_sgl_selection[
  order(sers_sgl_selection$wavenumber),
  ,
  drop = FALSE
]

sers_sgl_selection$window_start <- min(sers_wavenumbers_raw) +
  (sers_sgl_selection$group - 1) * sgl_window_width

sers_sgl_selection$window_end <- (
  sers_sgl_selection$window_start +
    sgl_window_width
)

sers_selected_regions <- aggregate(
  wavenumber ~ group + window_start + window_end,
  data = sers_sgl_selection,
  FUN = function(z) {
    paste(
      round(sort(z), 2),
      collapse = "; "
    )
  }
)

group_feature_counts <- table(
  sers_sgl_selection$group
)

sers_selected_regions$n_selected_wavenumbers <- as.integer(
  group_feature_counts[
    as.character(sers_selected_regions$group)
  ]
)

sers_selected_regions <- sers_selected_regions[
  order(sers_selected_regions$window_start),
  ,
  drop = FALSE
]

cat(
  "Selected spectral regions:",
  nrow(sers_selected_regions),
  "\n"
)

print(sers_selected_regions)

}

```

## Select spectral windows using a genetic algorithm

The GA searches among the same contiguous Raman windows used by SGL. Its fitness is the mean balanced accuracy from replicate-aware cross-validation minus a small penalty for the number of selected regions.

```{r sers-ga}
if (sers_selection_method == "GA") {

  set.seed(123)

  ga_min_regions <- 2L
  ga_max_regions <- min(10L, length(unique(sers_grouping)))
  ga_region_penalty <- 0.002

  # Represent every fixed Raman window by its mean intensity.
  all_group_ids <- sort(unique(sers_grouping))

  sers_window_matrix <- vapply(
    all_group_ids,
    function(group_id) {
      rowMeans(
        sers_x_raw[, sers_grouping == group_id, drop = FALSE],
        na.rm = TRUE
      )
    },
    numeric(nrow(sers_x_raw))
  )

  sers_window_matrix <- as.matrix(sers_window_matrix)
  colnames(sers_window_matrix) <- paste0("window_", all_group_ids)

  balanced_accuracy <- function(observed, predicted) {
    class_recall <- vapply(
      levels(sers_classes),
      function(class_name) {
        class_rows <- observed == class_name
        mean(predicted[class_rows] == class_name)
      },
      numeric(1)
    )

    mean(class_recall)
  }

  ga_fitness <- function(solution) {
    selected_windows <- which(solution == 1)
    n_selected <- length(selected_windows)

    if (
      n_selected < ga_min_regions ||
      n_selected > ga_max_regions
    ) {
      return(-1e6)
    }

    fold_scores <- vapply(
      sgl_cv_indices,
      function(test_rows) {
        train_rows <- setdiff(seq_len(nrow(sers_window_matrix)), test_rows)

        x_train <- sers_window_matrix[
          train_rows,
          selected_windows,
          drop = FALSE
        ]

        x_test <- sers_window_matrix[
          test_rows,
          selected_windows,
          drop = FALSE
        ]

        train_center <- colMeans(x_train)
        train_scale <- apply(x_train, 2, stats::sd)
        train_scale[!is.finite(train_scale) | train_scale == 0] <- 1

        x_train <- sweep(x_train, 2, train_center, "-")
        x_train <- sweep(x_train, 2, train_scale, "/")
        x_test <- sweep(x_test, 2, train_center, "-")
        x_test <- sweep(x_test, 2, train_scale, "/")

        class_centroids <- vapply(
          levels(sers_classes),
          function(class_name) {
            colMeans(
              x_train[
                sers_classes[train_rows] == class_name,
                ,
                drop = FALSE
              ]
            )
          },
          numeric(ncol(x_train))
        )

        class_centroids <- as.matrix(class_centroids)

        distances <- vapply(
          seq_len(ncol(class_centroids)),
          function(class_index) {
            rowSums(
              sweep(
                x_test,
                2,
                class_centroids[, class_index],
                "-"
              )^2
            )
          },
          numeric(nrow(x_test))
        )

        predicted <- factor(
          levels(sers_classes)[
            max.col(-distances, ties.method = "first")
          ],
          levels = levels(sers_classes)
        )

        balanced_accuracy(
          observed = sers_classes[test_rows],
          predicted = predicted
        )
      },
      numeric(1)
    )

    mean(fold_scores) - ga_region_penalty * n_selected
  }

  # Start with valid solutions containing 2-10 regions. This avoids
  # an initial population dominated by invalid large subsets.
  ga_initial_population <- matrix(
    0,
    nrow = 100,
    ncol = ncol(sers_window_matrix)
  )

  for (population_row in seq_len(nrow(ga_initial_population))) {
    initial_size <- sample(
      ga_min_regions:ga_max_regions,
      size = 1
    )

    ga_initial_population[
      population_row,
      sample(seq_len(ncol(sers_window_matrix)), initial_size)
    ] <- 1
  }

  ga_fit <- GA::ga(
    type = "binary",
    fitness = ga_fitness,
    nBits = ncol(sers_window_matrix),
    popSize = 100,
    maxiter = 300,
    run = 50,
    elitism = 5,
    pcrossover = 0.8,
    pmutation = 0.1,
    suggestions = ga_initial_population,
    seed = 123,
    monitor = FALSE
  )

  ga_solution <- ga_fit@solution

  if (is.matrix(ga_solution)) {
    ga_solution <- ga_solution[1, ]
  }

  selected_group_ids <- all_group_ids[which(ga_solution == 1)]

  if (
    length(selected_group_ids) < ga_min_regions ||
    length(selected_group_ids) > ga_max_regions
  ) {
    stop("The GA solution does not satisfy the region-count constraints.")
  }

  selected_idx <- which(sers_grouping %in% selected_group_ids)
  sers_selected_features <- colnames(sers_x_raw)[selected_idx]
  sers_active_grouping <- sers_grouping

  sers_sgl_selection <- data.frame(
    feature = sers_selected_features,
    wavenumber = sers_wavenumbers_raw[selected_idx],
    group = sers_grouping[selected_idx],
    stringsAsFactors = FALSE
  )

  sers_sgl_selection$window_start <- min(sers_wavenumbers_raw) +
    (sers_sgl_selection$group - 1) * sgl_window_width

  sers_sgl_selection$window_end <-
    sers_sgl_selection$window_start + sgl_window_width

  sers_selected_regions <- aggregate(
    wavenumber ~ group + window_start + window_end,
    data = sers_sgl_selection,
    FUN = function(z) paste(round(sort(z), 2), collapse = "; ")
  )

  sers_selected_regions$n_selected_wavenumbers <- as.integer(
    table(sers_sgl_selection$group)[
      as.character(sers_selected_regions$group)
    ]
  )

  sers_selected_regions <- sers_selected_regions[
    order(sers_selected_regions$window_start),
    ,
    drop = FALSE
  ]

  ga_summary <- data.frame(
    fitness = ga_fit@fitnessValue,
    mean_cv_balanced_accuracy = ga_fit@fitnessValue +
      ga_region_penalty * length(selected_group_ids),
    region_penalty = ga_region_penalty,
    selected_regions = length(selected_group_ids),
    stringsAsFactors = FALSE
  )

  cat("GA-selected spectral regions:", nrow(sers_selected_regions), "\n")
  print(sers_selected_regions)
  print(ga_summary)
}
```

## Construct variable-width windows using adjacent correlations

The WS method groups consecutive Raman shifts into the same window when their Pearson correlation across spectra is at least the selected threshold. Every variable-width window is represented by its mean intensity before aggregation to the six biological samples. This grouping is unsupervised and does not use the cell-line labels.

```{r sers-ws}
if (sers_selection_method == "WS") {

  adjacent_correlations <- vapply(
    seq_len(ncol(sers_x_raw) - 1L),
    function(index) {
      stats::cor(
        sers_x_raw[, index],
        sers_x_raw[, index + 1L],
        method = "pearson",
        use = "pairwise.complete.obs"
      )
    },
    numeric(1)
  )

  join_adjacent <- is.finite(adjacent_correlations) &
    adjacent_correlations >= ws_correlation_threshold

  # Start a new window whenever the adjacent correlation falls
  # below the selected threshold.
  ws_grouping <- cumsum(
    c(1L, as.integer(!join_adjacent))
  )

  if (length(ws_grouping) != ncol(sers_x_raw)) {
    stop("WS grouping does not match the number of SERS variables.")
  }

  if (length(unique(ws_grouping)) < 2L) {
    stop(
      paste(
        "WS produced fewer than two windows.",
        "Increase ws_correlation_threshold."
      )
    )
  }

  sers_selected_features <- colnames(sers_x_raw)
  sers_active_grouping <- ws_grouping
  selected_idx <- seq_len(ncol(sers_x_raw))

  ws_group_ids <- sort(unique(ws_grouping))

  ws_window_start <- vapply(
    ws_group_ids,
    function(group_id) {
      min(sers_wavenumbers_raw[ws_grouping == group_id])
    },
    numeric(1)
  )

  ws_window_end <- vapply(
    ws_group_ids,
    function(group_id) {
      max(sers_wavenumbers_raw[ws_grouping == group_id])
    },
    numeric(1)
  )

  sers_sgl_selection <- data.frame(
    feature = sers_selected_features,
    wavenumber = sers_wavenumbers_raw,
    group = ws_grouping,
    window_start = ws_window_start[
      match(ws_grouping, ws_group_ids)
    ],
    window_end = ws_window_end[
      match(ws_grouping, ws_group_ids)
    ],
    stringsAsFactors = FALSE
  )

  sers_selected_regions <- data.frame(
    group = ws_group_ids,
    window_start = ws_window_start,
    window_end = ws_window_end,
    wavenumber = vapply(
      ws_group_ids,
      function(group_id) {
        paste(
          round(
            sers_wavenumbers_raw[ws_grouping == group_id],
            2
          ),
          collapse = "; "
        )
      },
      character(1)
    ),
    n_selected_wavenumbers = as.integer(table(ws_grouping)),
    stringsAsFactors = FALSE
  )

  ws_adjacent_correlations <- data.frame(
    lower_wavenumber = sers_wavenumbers_raw[
      seq_len(length(sers_wavenumbers_raw) - 1L)
    ],
    upper_wavenumber = sers_wavenumbers_raw[
      seq.int(2L, length(sers_wavenumbers_raw))
    ],
    correlation = adjacent_correlations,
    joined_into_same_window = join_adjacent,
    stringsAsFactors = FALSE
  )

  ws_summary <- data.frame(
    correlation_threshold = ws_correlation_threshold,
    number_of_windows = length(ws_group_ids),
    minimum_window_variables = min(table(ws_grouping)),
    median_window_variables = stats::median(table(ws_grouping)),
    maximum_window_variables = max(table(ws_grouping)),
    stringsAsFactors = FALSE
  )

  cat("WS variable-width spectral regions:", nrow(sers_selected_regions), "\n")
  print(sers_selected_regions)
  print(ws_summary)
}
```

## Export the SERS selection results

```{r sers-export}
if (sers_selection_method == "SGL") {

write.table(
  sers_sgl_selection,
  file = result_file("selected_SERS_wavenumbers.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

write.table(
  sers_selected_regions,
  file = result_file("selected_SERS_regions.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

write.table(
  candidate_summary,
  file = result_file("tied_best_models.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

sgl_cv_summary <- data.frame(
  model_index = seq_along(sgl_error),
  cv_error = sgl_error,
  stringsAsFactors = FALSE
)

write.table(
  sgl_cv_summary,
  file = result_file("cross_validation_error_path.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

saveRDS(
  sgl_cv,
  result_file("cross_validation.rds")
)

saveRDS(
  sgl_fit,
  result_file("SERS_selection_model.rds")
)

}

if (sers_selection_method == "GA") {

  write.table(
    sers_sgl_selection,
    file = result_file("selected_SERS_wavenumbers.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  write.table(
    sers_selected_regions,
    file = result_file("selected_SERS_regions.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  write.table(
    ga_summary,
    file = result_file("selection_summary.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  saveRDS(
    ga_fit,
    result_file("SERS_selection_model.rds")
  )
}

if (sers_selection_method == "WS") {

  write.table(
    sers_sgl_selection,
    file = result_file("selected_SERS_wavenumbers.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  write.table(
    sers_selected_regions,
    file = result_file("selected_SERS_regions.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  write.table(
    ws_adjacent_correlations,
    file = result_file("adjacent_wavenumber_correlations.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  write.table(
    ws_summary,
    file = result_file("selection_summary.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

```

## Aggregate selected spectral regions by biological sample

```{r sers-aggregation}
selected_idx <- match(
  sers_selected_features,
  colnames(sers_x_raw)
)

selected_groups <- sers_active_grouping[selected_idx]

sers_x_selected <- sers_x_raw[
  ,
  sers_selected_features,
  drop = FALSE
]

# Average selected wavenumbers belonging to the same selected window
sers_region_matrix <- sapply(
  sort(unique(selected_groups)),
  function(group_id) {

    group_columns <- which(selected_groups == group_id)

    rowMeans(
      sers_x_selected[, group_columns, drop = FALSE],
      na.rm = TRUE
    )
  }
)

sers_region_matrix <- as.matrix(sers_region_matrix)

selected_group_ids <- sort(unique(selected_groups))

# Obtain the method-specific fixed or variable window boundaries.
region_match <- match(
  selected_group_ids,
  sers_selected_regions$group
)

if (anyNA(region_match)) {
  stop("Could not match selected groups to their spectral boundaries.")
}

region_start <- sers_selected_regions$window_start[region_match]
region_end <- sers_selected_regions$window_end[region_match]

colnames(sers_region_matrix) <- paste0(
  "SERS_region_",
  round(region_start, 1),
  "_",
  round(region_end, 1)
)

cat(
  paste("Individual", sers_selection_method, "selected wavenumbers:"),
  ncol(sers_x_selected),
  "\n"
)

cat(
  paste(sers_selection_method, "spectral regions after aggregation:"),
  ncol(sers_region_matrix),
  "\n"
)

# Aggregate spectra to six biological samples

sers_sum <- rowsum(
  sers_region_matrix,
  group = sers_sample_id,
  reorder = FALSE
)

sers_n <- table(sers_sample_id)

sers_mat <- sweep(
  sers_sum,
  MARGIN = 1,
  STATS = as.numeric(sers_n[rownames(sers_sum)]),
  FUN = "/"
)

sers_mat <- sers_mat[
  expected_samples,
  ,
  drop = FALSE
]

stopifnot(
  nrow(sers_mat) == 6,
  ncol(sers_mat) > 1,
  !anyNA(sers_mat),
  all(is.finite(sers_mat))
)

cat("Region-level SERS matrix supplied to DIABLO:\n")
print(dim(sers_mat))

print(colnames(sers_mat))
```

# # Match samples and define group information across all omics datasets.

```{r sample-info}
# Match samples across all omics datasets

common_samples <- Reduce(
  intersect,
  list(
    colnames(rna_counts),
    colnames(prot_mat),
    rownames(sers_mat)
  )
)

common_samples <- expected_samples[
  expected_samples %in% common_samples
]

stopifnot(identical(common_samples, expected_samples))

rna_counts <- rna_counts[, common_samples, drop = FALSE]
prot_mat <- prot_mat[, common_samples, drop = FALSE]
sers_mat <- sers_mat[common_samples, , drop = FALSE]

sample_info <- data.frame(
  sample_id = common_samples,
  group = factor(
    get_group(common_samples),
    levels = c("1G", "DLN")
  ),
  biological_replicate = factor(
    get_biorep(common_samples),
    levels = c("A", "B", "C")
  )
)

write.table(
  sample_info,
  result_file("sample_info_from_names.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

Y <- sample_info$group

stopifnot(
  length(levels(droplevels(Y))) == 2
)

sample_info
summary(Y)
```

## Filter and transform the RNA-seq data

```{r rna-normalise}
# Filter and variance-stabilize the RNA-seq data

keep_gene <- filterByExpr(
  rna_counts,
  group = Y
)

rna_counts_f <- rna_counts[
  keep_gene,
  ,
  drop = FALSE
]

dds <- DESeqDataSetFromMatrix(
  countData = rna_counts_f,
  colData = data.frame(
    group = Y,
    row.names = colnames(rna_counts_f)
  ),
  design = ~ group
)

dds <- estimateSizeFactors(dds)

rna_vst <- assay(
  vst(dds, blind = FALSE)
)

rna_block <- t(rna_vst)
rna_block <- as.matrix(rna_block)

stopifnot(
  ncol(rna_block) > 1,
  !anyNA(rna_block),
  all(is.finite(rna_block))
)

dim(rna_block)
rna_block[
  1:min(5, nrow(rna_block)),
  1:min(5, ncol(rna_block))
]
```

## Transform the proteomics data

```{r prot-normalise}
# Log2-transform the raw LFQ intensities

prot_block <- t(prot_mat)
prot_block <- as.matrix(prot_block)

colnames(prot_block) <- make.unique(
  colnames(prot_block)
)

stopifnot(
  ncol(prot_block) > 1,
  !anyNA(prot_block),
  all(is.finite(prot_block))
)

dim(prot_block)
prot_block[
  1:min(5, nrow(prot_block)),
  1:min(5, ncol(prot_block))
]
```

## Prepare the SERS data

```{r sers-normalise}
# Prepare the preprocessed SERS region matrix

sers_block <- as.matrix(sers_mat)
storage.mode(sers_block) <- "double"

colnames(sers_block) <- make.unique(
  colnames(sers_block)
)

stopifnot(
  ncol(sers_block) > 1,
  !anyNA(sers_block),
  all(is.finite(sers_block))
)

dim(sers_block)
sers_block[
  1:min(5, nrow(sers_block)),
  1:min(5, ncol(sers_block))
]
```

## Map RNA Ensembl identifiers to gene symbols

```{r gene-annotation}
# Map Ensembl identifiers to gene symbols

rna_feature_original <- colnames(rna_block)

rna_feature_clean <- trimws(
  sub("\\.[0-9]+$", "", rna_feature_original)
)

is_ensembl <- grepl(
  "^ENSG[0-9]+$",
  rna_feature_clean
)

gene_symbols <- rep(
  NA_character_,
  length(rna_feature_clean)
)

if (any(is_ensembl)) {
  gene_symbols[is_ensembl] <- as.character(
    AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys = rna_feature_clean[is_ensembl],
      keytype = "ENSEMBL",
      column = "SYMBOL",
      multiVals = "first"
    )
  )
}

rna_labels <- ifelse(
  !is.na(gene_symbols) & gene_symbols != "",
  gene_symbols,
  rna_feature_clean
)

colnames(rna_block) <- make.unique(rna_labels)

rna_annotation <- data.frame(
  original_feature = rna_feature_original,
  cleaned_feature = rna_feature_clean,
  is_ensembl = is_ensembl,
  gene_symbol = gene_symbols,
  diablo_label = colnames(rna_block),
  stringsAsFactors = FALSE
)

write.table(
  rna_annotation,
  result_file("DIABLO_RNA_gene_annotation.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

cat(
  "Total RNA features:",
  nrow(rna_annotation),
  "\n"
)

cat(
  "Features mapped to gene symbols:",
  sum(!is.na(gene_symbols)),
  "\n"
)

cat(
  "Features retained as Ensembl IDs:",
  sum(is.na(gene_symbols)),
  "\n"
)

head(rna_annotation)
```

## Feature pre-filtering and scaling

```{r prefilter}
# Select high-variance features before scaling

top_rna <- 2000
top_prot <- 1000
top_sers <- 1000

filter_and_select <- function(
  x,
  n_features,
  block_name
) {
  x <- as.matrix(x)
  storage.mode(x) <- "double"

  # Remove features containing non-finite values
  finite_features <- apply(
    x,
    2,
    function(z) all(is.finite(z))
  )

  cat(
    block_name,
    "non-finite features removed:",
    sum(!finite_features),
    "\n"
  )

  x <- x[, finite_features, drop = FALSE]

  # Calculate variance before scaling
  feature_variance <- apply(
    x,
    2,
    var
  )

  variable_features <- is.finite(feature_variance) &
    feature_variance > 0

  cat(
    block_name,
    "zero-variance features removed:",
    sum(!variable_features),
    "\n"
  )

  x <- x[, variable_features, drop = FALSE]
  feature_variance <- feature_variance[variable_features]

  stopifnot(ncol(x) > 1)

  # Retain features with the highest variance
  n_select <- min(
    n_features,
    ncol(x)
  )

  selected_features <- names(
    sort(
      feature_variance,
      decreasing = TRUE
    )
  )[seq_len(n_select)]

  x <- x[
    ,
    selected_features,
    drop = FALSE
  ]

  cat(
    block_name,
    "features retained:",
    ncol(x),
    "\n"
  )

  x
}

rna_block <- filter_and_select(
  rna_block,
  top_rna,
  "RNA"
)

prot_block <- filter_and_select(
  prot_block,
  top_prot,
  "Proteome"
)

sers_block <- filter_and_select(
  sers_block,
  top_sers,
  "SERS"
)

# Scale the selected features
rna_block <- as.matrix(
  scale(rna_block)
)

prot_block <- as.matrix(
  scale(prot_block)
)

sers_block <- as.matrix(
  scale(sers_block)
)

# Ensure unique feature labels
colnames(rna_block) <- make.unique(
  colnames(rna_block)
)

colnames(prot_block) <- make.unique(
  colnames(prot_block)
)

colnames(sers_block) <- make.unique(
  colnames(sers_block)
)

# Verify sample correspondence
stopifnot(
  identical(
    rownames(rna_block),
    expected_samples
  ),
  identical(
    rownames(prot_block),
    expected_samples
  ),
  identical(
    rownames(sers_block),
    expected_samples
  )
)

# Verify that all matrices are suitable for DIABLO
stopifnot(
  all(is.finite(rna_block)),
  all(is.finite(prot_block)),
  all(is.finite(sers_block)),
  nrow(rna_block) == length(Y),
  nrow(prot_block) == length(Y),
  nrow(sers_block) == length(Y),
  ncol(rna_block) > 1,
  ncol(prot_block) > 1,
  ncol(sers_block) > 1
)

data <- list(
  transcriptome = rna_block,
  proteome = prot_block,
  sers = sers_block
)

lapply(data, dim)
summary(Y)
```

# Pairwise sPLS comparisons between all omics blocks

```{r pairwise-pls, fig.width=8, fig.height=6}
sers_pairwise_keep <- rep(
  min(9L, ncol(data[["sers"]])),
  2
)

pls_rna_prot <- spls(
  X = data[["transcriptome"]],
  Y = data[["proteome"]],
  ncomp = 2,
  keepX = c(25, 25),
  keepY = c(25, 25)
)

plotVar(
  pls_rna_prot,
  cutoff = 0.5,
  title = "Transcriptome vs proteome",
  legend = c("Transcriptome", "Proteome"),
  var.names = FALSE,
  style = "graphics",
  pch = c(16, 17),
  cex = c(2, 2),
  col = c("brown1", "lightgreen")
)

pls_rna_sers <- spls(
  X = data[["transcriptome"]],
  Y = data[["sers"]],
  ncomp = 2,
  keepX = c(25, 25),
  keepY = sers_pairwise_keep
)

plotVar(
  pls_rna_sers,
  cutoff = 0.5,
  title = "Transcriptome vs SERS",
  legend = c("Transcriptome", "SERS"),
  var.names = FALSE,
  style = "graphics",
  pch = c(16, 18),
  cex = c(2, 2),
  col = c("brown1", "steelblue")
)

pls_prot_sers <- spls(
  X = data[["proteome"]],
  Y = data[["sers"]],
  ncomp = 2,
  keepX = c(25, 25),
  keepY = sers_pairwise_keep
)

plotVar(
  pls_prot_sers,
  cutoff = 0.5,
  title = "Proteome vs SERS",
  legend = c("Proteome", "SERS"),
  var.names = FALSE,
  style = "graphics",
  pch = c(17, 18),
  cex = c(2, 2),
  col = c("lightgreen", "steelblue")
)
```
## Correlations between pairwise sPLS components

```{r pairwise-correlations}
cor_rna_prot <- cor(
  pls_rna_prot$variates$X,
  pls_rna_prot$variates$Y
)

cor_rna_sers <- cor(
  pls_rna_sers$variates$X,
  pls_rna_sers$variates$Y
)

cor_prot_sers <- cor(
  pls_prot_sers$variates$X,
  pls_prot_sers$variates$Y
)

pairwise_component_correlations <- data.frame(
  comparison = c(
    "Transcriptome–proteome",
    "Transcriptome–SERS",
    "Proteome–SERS"
  ),
  component_1 = c(
    cor_rna_prot[1, 1],
    cor_rna_sers[1, 1],
    cor_prot_sers[1, 1]
  ),
  component_2 = c(
    cor_rna_prot[2, 2],
    cor_rna_sers[2, 2],
    cor_prot_sers[2, 2]
  )
)

pairwise_component_correlations
```

# DIABLO model
## Define the fully connected DIABLO design matrix

```{r diablo-design}
design <- matrix(
  1,
  nrow = length(data),
  ncol = length(data),
  dimnames = list(
    names(data),
    names(data)
  )
)

diag(design) <- 0

design
```

## Fit the initial DIABLO model

```{r diablo}
ncomp_start <- 2

set.seed(123)

basic_diablo_model <- block.splsda(
  X = data,
  Y = Y,
  ncomp = ncomp_start,
  design = design,
  near.zero.var = TRUE
)

basic_diablo_model
```

## Evaluate the number of components

```{r num-comp}
n_folds <- min(
  3,
  min(table(Y))
)

set.seed(123)

perf_initial <- perf(
  basic_diablo_model,
  validation = "Mfold",
  folds = n_folds,
  nrepeat = 10,
  dist = "centroids.dist",
  BPPARAM = BiocParallel::SerialParam()
)

plot(
  perf_initial,
  sd = TRUE
)

perf_initial$error.rate
```
## Select the optimal number of components

```{r select-ncomp}
ncomp_tuned <- perf_initial$choice.ncomp$WeightedVote[
  "Overall.BER",
  "centroids.dist"
]

ncomp_final <- as.numeric(ncomp_tuned)

stopifnot(
  is.finite(ncomp_final),
  ncomp_final >= 1,
  ncomp_final <= ncomp_start
)

cat(
  "Selected number of components:",
  ncomp_final,
  "\n"
)

perf_initial$choice.ncomp$WeightedVote
```

## Tune the number of selected features

```{r tune-features}
test_keep_x <- list(
  transcriptome = c(5, 10, 15),
  proteome = c(5, 10, 15),
  sers = c(2, 5, 9)
)

# Ensure that candidates do not exceed the available features
test_keep_x <- Map(
  function(candidates, block) {
    candidates[candidates <= ncol(block)]
  },
  candidates = test_keep_x,
  block = data
)

set.seed(123)

tune_diablo <- tune.block.splsda(
  X = data,
  Y = Y,
  ncomp = ncomp_final,
  test.keepX = test_keep_x,
  design = design,
  validation = "Mfold",
  folds = n_folds,
  nrepeat = 10,
  dist = "centroids.dist",
  near.zero.var = TRUE,
  BPPARAM = BiocParallel::SerialParam()
)

tune_diablo$choice.keepX
```
## Extract the tuned feature numbers

```{r extract-keepx}
keep_x_final <- tune_diablo$choice.keepX

keep_x_final
```

# Final model
## Fit the final sparse DIABLO model

```{r diablo-final}
set.seed(123)

final_diablo_model <- block.splsda(
  X = data,
  Y = Y,
  ncomp = ncomp_final,
  keepX = keep_x_final,
  design = design,
  near.zero.var = TRUE
)

saveRDS(
  final_diablo_model,
  result_file("DIABLO_final_model.rds")
)

final_diablo_model
```

## Display the design matrix used in the final model

```{r diablo-2}
final_diablo_model$design
```

## Extract and export the selected DIABLO features

```{r selected-features}
selected_features <- lapply(
  names(data),
  function(block_name) {
    selected <- selectVar(
      final_diablo_model,
      block = block_name,
      comp = 1
    )[[block_name]]

    data.frame(
      block = block_name,
      component = 1,
      feature = selected$name,
      loading = as.numeric(selected$value[, 1]),
      stringsAsFactors = FALSE
    )
  }
)

selected_features <- bind_rows(
  selected_features
)

write.table(
  selected_features,
  result_file("DIABLO_selected_features.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

selected_features
```

## Displaying DIABLO loading weights

```{r plot-helper-functions}
short_label <- function(x, max_chars = 32) {
  x <- as.character(x)

  ifelse(
    nchar(x) > max_chars,
    paste0(
      substr(x, 1, max_chars - 3),
      "..."
    ),
    x
  )
}

plot_block_loadings <- function(
  model,
  block,
  comp = 1,
  top_n = 15,
  main = NULL
) {
  loadings <- model$loadings[[block]][, comp]

  loadings <- loadings[
    is.finite(loadings) & loadings != 0
  ]

  if (length(loadings) == 0) {
    stop(
      "No non-zero loadings were found for ",
      block,
      " on component ",
      comp,
      "."
    )
  }

  loadings <- loadings[
    order(
      abs(loadings),
      decreasing = TRUE
    )
  ]

  loadings <- head(
    loadings,
    top_n
  )

  loadings <- rev(loadings)

  names(loadings) <- short_label(
    names(loadings)
  )

  loading_colours <- ifelse(
    loadings > 0,
    "darkorange2",
    "dodgerblue3"
  )

  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)

  par(
    mar = c(5, 12, 4, 2),
    cex = 0.85
  )

  barplot(
    loadings,
    horiz = TRUE,
    las = 1,
    cex.names = 1,
    col = loading_colours,
    border = "grey25",
    main = if (is.null(main)) {
      paste(
        block,
        "component",
        comp,
        "loadings"
      )
    } else {
      main
    },
    xlab = "Loading weight"
  )

  abline(
    v = 0,
    lty = 2,
    lwd = 1.5
  )

  legend(
    "bottomright",
    legend = c(
      "Positive loading",
      "Negative loading"
    ),
    fill = c(
      "darkorange2",
      "dodgerblue3"
    ),
    border = "grey25",
    bty = "n",
    cex = 0.9
  )
}
```

## Plot loadings from the single selected component

```{r loading-plots}
plot_block_loadings(
  final_diablo_model,
  block = "transcriptome",
  comp = 1,
  main = "Transcriptome loadings"
)

plot_block_loadings(
  final_diablo_model,
  block = "proteome",
  comp = 1,
  main = "Proteome loadings"
)

plot_block_loadings(
  final_diablo_model,
  block = "sers",
  comp = 1,
  main = "SERS loadings"
)
```

# DIABLO visualizations

```{r visualization-model}
# Fit a two-component model for exploratory visualization

set.seed(123)

visualization_model <- block.splsda(
  X = data,
  Y = Y,
  ncomp = 2,
  keepX = list(
    transcriptome = c(5, 5),
    proteome = c(5, 5),
    sers = c(2, 2)
  ),
  design = design,
  near.zero.var = TRUE
)
```

```{r plot-diablo}
plotDiablo(
  visualization_model,
  ncomp = 2
)
```

```{r plot-individuals}
plotIndiv(
  visualization_model,
  comp = c(1, 2),
  ind.names = FALSE,
  legend = TRUE,
  title = "DIABLO sample plot"
)
```

```{r plot-arrows}
plotArrow(
  visualization_model,
  comp = c(1, 2),
  ind.names = FALSE,
  legend = TRUE,
  title = "DIABLO multi-omics sample agreement"
)
```

```{r plot-variables}
plotVar(
  visualization_model,
  comp = c(1, 2),
  var.names = FALSE,
  style = "graphics",
  legend = TRUE,
  pch = c(16, 17, 15),
  cex = c(2, 2, 2),
  col = c(
    "brown1",
    "lightgreen",
    "steelblue"
  )
)
```

# Correlation plots

```{r circos-plot, fig.width=20, fig.height=16, out.width="100%"}
circosPlot(
  final_diablo_model,
  cutoff = 0.7,
  line = TRUE,
  color.blocks = c(
    "brown1",
    "lightgreen",
    "steelblue"
  ),
  color.cor = c(
    "chocolate3",
    "grey20"
  ),
  size.labels = 1.35,
  size.variables = 1.35,
  font.labels = 2,
  font = 2,
  block.line = 3,
  lwd = 2
)
```

```{r network-plot, fig.width=14, fig.height=10, out.width="90%"}
set.seed(123)

mixOmics::network(
  final_diablo_model,
  blocks = c(1, 2, 3),
  color.node = c(
    "brown1",
    "lightgreen",
    "steelblue"
  ),
  cutoff = 0.4,
  size.node = 0.55
)
```

# Clustered image map

```{r cim-diablo, fig.width=24, fig.height=14, out.width="100%"}
make_labelled_diablo_heatmap <- function(model) {
  block_label_map <- c(
    transcriptome = "Gene",
    proteome = "Protein",
    sers = "SERS"
  )

  selected_matrices <- lapply(names(model$X), function(block_name) {
    block_loadings <- as.matrix(model$loadings[[block_name]])
    selected <- rowSums(abs(block_loadings), na.rm = TRUE) > 0
    model$X[[block_name]][, selected, drop = FALSE]
  })

  names(selected_matrices) <- names(model$X)
  heatmap_matrix <- do.call(cbind, selected_matrices)

  variable_blocks <- unlist(
    Map(
      function(block_name, block_matrix) {
        rep(block_label_map[[block_name]], ncol(block_matrix))
      },
      names(selected_matrices),
      selected_matrices
    ),
    use.names = FALSE
  )

  sample_groups <- ifelse(
    as.character(model$Y) == "1G",
    "MCM-1G",
    "MCM-DLN"
  )

  sample_groups <- factor(
    sample_groups,
    levels = c("MCM-DLN", "MCM-1G")
  )

  variable_blocks <- factor(
    variable_blocks,
    levels = c("Gene", "Protein", "SERS")
  )

  sample_order <- order(sample_groups)
  heatmap_matrix <- heatmap_matrix[sample_order, , drop = FALSE]
  sample_groups <- sample_groups[sample_order]

  # Match cimDiablo's display range while retaining the selected values
  heatmap_matrix[heatmap_matrix > 2] <- 2
  heatmap_matrix[heatmap_matrix < -2] <- -2

  # Write the condition names directly into the coloured left blocks
  group_annotation <- ComplexHeatmap::rowAnnotation(
    Group = ComplexHeatmap::anno_block(
      gp = grid::gpar(fill = c("#F28E2B", "#2C7FB8"), col = "white"),
      labels = c("MCM-DLN", "MCM-1G"),
      labels_gp = grid::gpar(
        col = "white",
        fontsize = 11,
        fontface = "bold"
      ),
      width = grid::unit(18, "mm")
    ),
    show_annotation_name = FALSE
  )

  # Write Gene, Protein and SERS directly into the coloured top blocks
  block_annotation <- ComplexHeatmap::HeatmapAnnotation(
    Block = ComplexHeatmap::anno_block(
      gp = grid::gpar(
        fill = c("#C44E52", "#55A868", "#4C72B0"),
        col = "white"
      ),
      labels = c("Gene", "Protein", "SERS"),
      labels_gp = grid::gpar(
        col = "white",
        fontsize = 11,
        fontface = "bold"
      ),
      height = grid::unit(9, "mm")
    ),
    show_annotation_name = FALSE
  )

  color_limit <- max(abs(heatmap_matrix), na.rm = TRUE)
  heatmap_colors <- circlize::colorRamp2(
    seq(-color_limit, color_limit, length.out = 64),
    mixOmics::color.jet(64)
  )

  ComplexHeatmap::Heatmap(
    heatmap_matrix,
    name = "Scaled value",
    col = heatmap_colors,
    row_split = sample_groups,
    column_split = variable_blocks,
    cluster_rows = FALSE,
    cluster_row_slices = FALSE,
    cluster_columns = TRUE,
    cluster_column_slices = FALSE,
    left_annotation = group_annotation,
    top_annotation = block_annotation,
    show_row_names = FALSE,
    show_column_names = TRUE,
    row_title = NULL,
    column_names_gp = grid::gpar(fontsize = 11, fontface = "bold"),
    column_names_rot = 45,
    column_names_max_height = grid::unit(85, "mm"),
    column_title = NULL,
    row_gap = grid::unit(8, "mm"),
    column_gap = grid::unit(3, "mm"),
    rect_gp = grid::gpar(col = "white", lwd = 0.6),
    border = TRUE,
    heatmap_legend_param = list(
      title = "Scaled value",
      legend_height = grid::unit(45, "mm")
    )
  )
}

labelled_diablo_heatmap <- make_labelled_diablo_heatmap(
  final_diablo_model
)

ComplexHeatmap::draw(
  labelled_diablo_heatmap,
  heatmap_legend_side = "right"
)
```

# Save figures

```{r save-figures}
save_png <- function(
  filename,
  plot_code,
  width = 3000,
  height = 2200,
  resolution = 250
) {
  png(
    filename = file.path(
      figure_dir,
      paste0(sers_method_tag, "_", filename, ".png")
    ),
    width = width,
    height = height,
    res = resolution
  )

  on.exit(dev.off(), add = TRUE)

  eval(
    substitute(plot_code),
    envir = parent.frame()
  )
}

save_png(
  "DIABLO_plot",
  plotDiablo(
    visualization_model,
    ncomp = 2
  )
)

save_png(
  "DIABLO_samples",
  plotIndiv(
    visualization_model,
    comp = c(1, 2),
    ind.names = FALSE,
    legend = TRUE,
    title = "DIABLO sample plot"
  )
)

save_png(
  "DIABLO_arrows",
  plotArrow(
    visualization_model,
    comp = c(1, 2),
    ind.names = FALSE,
    legend = TRUE,
    title = "DIABLO multi-omics sample agreement"
  )
)

save_png(
  "DIABLO_variables",
  plotVar(
    visualization_model,
    comp = c(1, 2),
    var.names = FALSE,
    style = "graphics",
    legend = TRUE,
    pch = c(16, 17, 15),
    cex = c(2, 2, 2),
    col = c(
      "brown1",
      "lightgreen",
      "steelblue"
    )
  )
)

save_png(
  "DIABLO_circos",
  circosPlot(
    final_diablo_model,
    cutoff = 0.7,
    line = TRUE,
    color.blocks = c(
      "brown1",
      "lightgreen",
      "steelblue"
    ),
    color.cor = c(
      "chocolate3",
      "grey20"
    ),
    size.labels = 1.35,
    size.variables = 1.35,
    font.labels = 2,
    font = 2,
    block.line = 3,
    lwd = 2
  ),
  width = 5000,
  height = 4000,
  resolution = 300
)


save_png(
  "DIABLO_transcriptome_loadings",
  plot_block_loadings(
    final_diablo_model,
    block = "transcriptome",
    comp = 1,
    top_n = 15,
    main = "Transcriptome loadings, component 1"
  )
)

save_png(
  "DIABLO_proteome_loadings",
  plot_block_loadings(
    final_diablo_model,
    block = "proteome",
    comp = 1,
    top_n = 15,
    main = "Proteome loadings, component 1"
  )
)

save_png(
  "DIABLO_SERS_loadings",
  plot_block_loadings(
    final_diablo_model,
    block = "sers",
    comp = 1,
    top_n = 15,
    main = "SERS loadings, component 1"
  )
)

save_png(
  "DIABLO_clustered_image_map",
  ComplexHeatmap::draw(
    labelled_diablo_heatmap,
    heatmap_legend_side = "right"
  ),
  width = 7800,
  height = 4800,
  resolution = 300
)

cat(
  "Figures saved in:",
  figure_dir,
  "\n"
)
```

```{r save-network-directly, sane_par=FALSE}
network_file <- file.path(
  figure_dir,
  paste0(sers_method_tag, "_DIABLO_network.png")
)

network_candidates <- list.files(
  path = knitr::opts_chunk$get("fig.path"),
  pattern = "^network-plot-[0-9]+[.]png$",
  full.names = TRUE
)

if (length(network_candidates) == 0L) {
  stop("Rendered network figure not found.")
}

network_source <- network_candidates[
  which.max(file.info(network_candidates)$mtime)
]

file.copy(network_source, network_file, overwrite = TRUE)

cat("Network saved to:", network_file, "\n")
```

```{r session-info}
sessionInfo()
```
